# The dual evolutionary foundations of political ideology^[Chapter based on "Claessens, S., Fischer, K., Chaudhuri, A., Sibley, C. G., & Atkinson, Q. D. (2020). The dual evolutionary foundations of political ideology. _Nature Human Behaviour_, _4_(4), 336–345" and "Claessens, S., Chaudhuri, A., Sibley, C. G., & Atkinson, Q. D. (in press). The evolutionary basis of political ideology. In C. G. Sibley & D. Osborne (Eds.), _Cambridge Handbook of Political Psychology_. Cambridge, UK: Cambridge University Press."]

## Introduction

In recent decades, the concept of political ideology has enjoyed a resurgence in the social sciences [@Jost2009]. Political ideology is defined as a set of stable, interrelated beliefs and attitudes that organise views on political and social issues. While scholars had previously attributed only a minor role to ideology in shaping political behaviour [@Converse1964; @Zaller1992], it has since become clear that political ideology both motivates voting and coherently structures views on a wide range of social issues, from taxation and welfare to crime and religion [@Jost2006]. Traditionally, ideology has been conceptualised as varying along a unidimensional spectrum, with liberalism on the left and conservatism on the right [@Adorno1950]. Broadly, liberalism emphasises equality, social change, and system reform, while conservatism emphasises hierarchy, conventionalism, and tradition. This left-right distinction dates back more than 200 years to the 1791 French legislative assembly (monarchists sat on the right) but remains the primary means of describing political opinion in social science and public discourse (Figure \@ref(fig:ch2figUniMulti)).

(ref:ch2figUniMultiCaption) _The number of scholarly and newspaper articles mentioning unidimensional and multidimensional approaches to political ideology._ Data are from three sources: the peer-reviewed literature database Scopus and two US newspapers, the Wall Street Journal and the New York Times. The review of Scopus selected the top 100 most-cited articles from the last 20 years under the search term 'political ideology'. The review of the US newspapers selected the top five most relevant articles in every year from 1999-2018 under the search term 'political ideology'. If the articles contained explicit references to unidimensionality or any of the terms liberal, conservative, Democrat, Republican, or variations thereupon, they were coded as mentioning unidimensionality. If the articles contained explicit references to multidimensionality or any of the terms social dominance, authoritarianism, economic conservatism, social conservatism, or variations thereupon, they were coded as mentioning multidimensionality. Figure adapted from @Claessens2020a.

```{r ch2figUniMulti, fig.cap="(ref:ch2figUniMultiCaption)", fig.align="center", fig.scap="The number of scholarly and newspaper articles mentioning unidimensional and multidimensional approaches to political ideology.", echo=F, fig.width=5, fig.height=5}
readd(ch2_figUniMulti)
```

Despite the popularity of this unidimensional model, political views cannot be neatly summarised by a single liberal-conservative spectrum [@Treier2009]. The 2016 US Presidential Election highlighted how divergent political views can be within left or right discourse, with disagreements between Hillary Clinton and Bernie Sanders within the Democratic Party and opposition to Donald Trump from within the Republican Party. In the electorate itself, many people express conflicting political beliefs that cross party lines [@Claassen2015]. Libertarians are a classic case of this misalignment, harbouring 'liberal' views on social issues but 'conservative' views regarding economic policy. It is perhaps not a surprise, then, that unidimensional self-report scales of political ideology often have low internal consistency [@Altemeyer1981] and low external validity [@Treier2009], and they frequently produce more than one latent variable in factor analyses [@Duckitt2009]. In short, a single left-right dimension misses important features of the political landscape.

By contrast, scholars from many disciplines have converged upon two dimensions of political ideology. These dual dimensions have repeatedly emerged in the literature over the last 50 years (Table \@ref(tab:ch2tabScales)), despite researchers using different methodologies and, indeed, different labels to capture ideology. Some researchers have focused on the attitudes that people hold about political and social issues, clustering these into correlated categories using data-driven, atheoretical factor-analytic methods [@Feldman2014; @Jost2003]. Others have defined core universal human values (e.g., benevolence, tradition, security) and then determined how they influence ideology [@Schwartz1992]. Lexical approaches have abstracted even further, allowing the underlying structure of political attitudes to emerge from ratings of dictionary-based 'isms' [e.g., Machiavellianism or traditionalism; @Saucier2000]. Moral psychology has inductively derived clusters of moral values and noted how they strongly predict political ideology [@Graham2009], and cross-cultural approaches have validated scale items across many different societies, finding that the same dimensions recur [@Ashton2005]. Across this myriad of methodologies, both exploratory and confirmatory, researchers have found very similar two-dimensional ideological structures, strongly suggesting that the scales in Table \@ref(tab:ch2tabScales) are all capturing the same underlying psychological phenomena.

(ref:ch2ref01) @Hughes1975

(ref:ch2ref02) @Duckitt2009

(ref:ch2ref03) @Eysenck1954

(ref:ch2ref04) @Tomkins1964

(ref:ch2ref05) @Rokeach1973

(ref:ch2ref06) @Hofstede1980

(ref:ch2ref07) @Kerlinger1984

(ref:ch2ref08) @Forsyth1980

(ref:ch2ref09) @Katz1988

(ref:ch2ref10) @Middendorp1991

(ref:ch2ref11) @Trompenaars1993

(ref:ch2ref12) @Braithwaite1994

(ref:ch2ref13) @Schwartz1996

(ref:ch2ref14) @Triandis1998

(ref:ch2ref15) @Saucier2000

(ref:ch2ref16) @Ashton2005

(ref:ch2ref17) @Stangor2006

(ref:ch2ref18) @Ferguson1939

(ref:ch2ref19) @Boski1993

(ref:ch2ref20) @Jost2003

(ref:ch2ref21) @Graham2009

(ref:ch2tableCaption) _Various labels for the two dimensions of political ideology._ Table adapted from @Duckitt2009 and @Claessens2020a.

```{r ch2tabScales, echo=F}
knitr::kable(
    readd(ch2_tabScales), booktabs = TRUE,
    format = "latex",
    caption = "(ref:ch2tableCaption)",
    caption.short = "Various definitions for the two dimensions of political ideology",
    col.names = gsub("[.]", " ", names(readd(ch2_tabScales)))
  ) %>%
  kable_styling(font_size = 10) %>%
  column_spec(1:3, width = "12em")
```

How should we understand these two dimensions of political ideology? The first dimension, often referred to as economic conservatism or social dominance, predicts stances on issues like taxation, government-funded healthcare, welfare programs, and free education [@Feldman2014]. Economic conservatives view the world as a 'competitive jungle', in which dominance, inequality, and power imbalances are commonplace [@Duckitt2009]. The second dimension, often referred to as social conservatism or authoritarianism, predicts stances on issues like traditional social values, criminal justice, patriotism, national security, same-sex marriage, and religion [@Altemeyer1981; @Feldman2014]. These social conservatives view the world as more threatening, dangerous, and unpredictable [@Duckitt2009].

It remains unclear why political attitudes tend to be structured along these two particular ideological dimensions and why the dual dimensions are linked with distinct worldviews. In this chapter, we argue that an evolutionary approach to political ideology can shed light on both of these questions. Recent research challenges the common assumption that ideological variation is best understood as the result of historically contingent social, cultural, and environmental factors [@Altemeyer1981; @Campbell1960; @Converse1964; @Zaller1992] with little basis in biology. Variation in political ideology is heritable [@Alford2004; @Alford2005; @Batricevic2017; @Hatemi2011; @Hatemi2007; @Kandler2016; @Lewis2014; @Verhulst2012], remains stable over long periods of time [@Block2006], and covaries with basic physiological differences [@Dodd2012; @Fodor2008; @Hibbing2014; @Oosterhoff2018; @Oxley2008; @Shook2009; @Smith2011; but see @Bakker2020; @Fiagbenu2021; @Osmundsen2019]. The two dimensions of ideology are also repeatedly observed across a wide range of cultures [@Ashton2005; @Duriez2005; @Pratto2013], suggesting that they may be universal. This recurrent pattern of ideological variation across cultures, together with heritable stable individual differences, points to the possibility that the two dimensions are at least partly grounded in biology. As such, an evolutionary approach can help us understand the dual-dimensional structure of political ideology.

## An evolutionary approach to psychology

Unlike other approaches to psychology that describe _how_ the human mind functions, an evolutionary approach to psychology also considers _why_ the human mind functions the way it does [@Scott-Phillips2011; @Tinbergen1963] by placing human psychology within the context of our species' particular evolutionary history. This approach acknowledges that many of the psychological mechanisms underlying modern human behaviour were inherited from our hominid ancestors [@Barkow1992]. Following the logic of evolution by natural selection [@Darwin1859], these psychological mechanisms are thought to be universal features of the human mind that are adapted to the basic challenges of survival and reproduction that have faced our lineage. For example, incest avoidance is hypothesised to be a psychological mechanism shared by all humans that functions to avoid the deleterious genetic consequences of inbreeding [@Westermarck1891].

As well as explaining universal features of the human mind, an evolutionary approach also explains variation in psychology and behaviour [@Buss2009]. First, heritable variation in psychological traits can be maintained by evolution if different levels of the trait are adaptive in different contexts; a mechanism known as balancing selection on fitness trade-offs [@Buss2009; @Nettle2006]. For example, the extraversion dimension of personality is hypothesised to result from the trade-offs underlying different extraversion levels [e.g., extraverts benefit from increased sexual encounters but also suffer from increased risk of illness; @Nettle2006]. Second, different environmental inputs into a universal psychological architecture can produce variation through phenotypic plasticity [@Snell-Rood2013]. Phenotypic plasticity adaptively calibrates psychology and behaviour to local ecological challenges [@Sng2018]. For example, the tendency for people lower in socio-economic status to discount future rewards is hypothesised to be an adaptive response to environmental harshness [@Pepper2017]. An evolutionary approach therefore explains diversity, as well as uniformity, in human psychology and behaviour.

Evolutionary accounts of human behaviour also emphasise that culture has been a key driving force in human evolution [@Henrich2015; @Richerson2005]. Unlike other primates, humans depend upon a large body of socially-learned beliefs, norms, and values in order to survive in novel environments. These cultural traits are inherited and accumulated within social groups over many generations, providing solutions to local adaptive problems (e.g., foraging, hunting, food processing) that individuals could not solve through trial-and-error alone [@Henrich2003]. As a result, humans have evolved to be keenly attentive to cultural information, readily learning social norms from a young age [@Rakoczy2008]. Humans also preferentially direct their learning towards individuals within the same cultural group, as delineated by cultural markers like accent [@Kinzler2009] or language [@Kinzler2007]. Thus, an evolutionary approach also highlights cultural variability [@Henrich2015], highly interdependent sociality [@Tomasello2012], and groupishness [@Bernhard2006] as key features of human psychology.

Daniel Dennett once described evolutionary theory as a universal acid which revolutionises every discipline it touches [@Dennett1995]. Indeed, evolutionary perspectives on human psychology continue to shed light on various topics, including personality [@Buss2009], culture [@Henrich2015], religion [@Norenzayan2016], and cooperation [@Nowak2006]. We will now review several recent evolutionary approaches to the two dimensions of political ideology.

## Evolutionary approaches to the two dimensions of political ideology

A number of promising evolutionary approaches have already
derived important insights about the economic and social dimensions of political ideology captured in Table \@ref(tab:ch2tabScales). 

A growing body of work shows a positive relationship between economic conservatism and physical dominance attributes [@Petersen2019; @Price2011; @Price2017]. One recent study found that, across twelve high-powered cross-cultural conceptual replications, both self-reported measures of physical formidability (e.g., motivation to build muscularity) and objective measures of formidability (e.g., handgrip strength, upper-body strength) correlated positively with economic conservatism and social dominance orientation in males, but not females [@Petersen2019]. In both our species and our primate cousins, conflicts over resources are often decided by differences in physical formidability, suggesting that the economic dimension of ideology may be related to resource conflict and competition.

Researchers have consistently found positive relationships between social conservatism and disgust sensitivity, pathogen avoidance, and parasite stress [@Hodson2007; @Inbar2009; @Inbar2012; @Tybur2010; for a meta-analysis, see @Terrizzi2013]. One large-scale cross-cultural study found that individuals higher in traditionalism, a measure of social conservatism, exhibited greater disgust sensitivity and were more likely to come from nations with higher parasite stress [@Tybur2016]. No relationship was found for economic conservatism. Based on these findings, scholars have suggested that social conservatism may reflect variation in the sensitivity of the behavioural immune system, a cluster of evolved behavioural and psychological mechanisms that provide organisms with a first line of defence against disease-causing microorganisms [@Schaller2011].

Other authors argue that the relationship between disgust sensitivity and social conservatism can be understood under the more general umbrella of threat sensitivity, or negativity bias, which they claim uniquely describes the core of social conservatism [@Hibbing2014]. A human cognitive bias increasing the salience of threatening stimuli makes evolutionary sense under the logic of error management theory [@Haselton2006], which posits that decisions made under uncertainty should be adaptively biased towards making less costly errors. Indeed, sensitivity to threatening stimuli that are _not_ disgusting, such as sudden noises [@Oxley2008], has also been linked to social conservatism. Under this view, social conservatives are self-protective and risk averse in their policy views (e.g., adhering to tried-and-tested social norms, opposing social change) because they have a stronger negativity bias and, thus, are acting to reduce the probability of negative events occurring.

Moral foundations theory [@Graham2009; @Haidt2012] provides further insight into the evolutionary foundations of the two dimensions of political ideology. This theory argues that moral reasoning can largely be explained by people's sensitivities to five evolved moral foundations: care-harm, fairness-cheating, loyalty-betrayal, authority-subversion, and sanctity-degradation. These foundations are hypothesised to be solutions to adaptive problems faced by humans in their evolutionary history. The care-harm foundation is related to adaptive problems of nurturing and protecting vulnerable individuals, such as infants, and the fairness-cheating foundation reflects problems of cooperating with others to mutual benefit and detecting cheaters. The loyalty-betrayal foundation relates to problems of coalition-formation and group defence, while the authority-subversion foundation reflects the need for an established authority to promote group coordination. The sanctity-degradation foundation responds to problems of disease and contamination.

These five moral foundations correlate with political ideology. Initial research focused on the liberal-conservative spectrum, showing that liberals advocate the care and fairness dimensions above the others, while conservatives give equal weight to all five dimensions [@Graham2009]. More recent work has begun to link the moral foundations to the two dimensions of ideology separately. Factor analysis shows that the five moral foundations can be reduced to two distinct sets: individualising foundations (care-harm, fairness-cheating) and binding foundations [loyalty-betrayal, authority-subversion, sanctity-degradation; @Graham2009; @Sinn2017]. @Federico2013 found that sensitivity to the individualising foundations correlates with social dominance orientation, whereas sensitivity to the binding foundations correlates with right-wing authoritarianism. Another study found that factor analysis of self-report items capturing the moral foundations and political ideology produced a two-factor solution: one with high loadings from individualising foundations and social dominance orientation, and the other with high loadings from binding foundations and right-wing authoritarianism [@Sinn2017]. These results suggest that, despite its proposed five factor structure, moral foundations theory has independently converged upon the same two dimensions of ideology that have been repeatedly identified in political psychology [@Duckitt2009]. The first (economic) dimension reflects adaptive problems relating to care, nurturance, and cooperation, and the second (social) dimension reflects problems relating to group defence, group coordination, and disease avoidance.

In sum, evolutionary approaches have provided important insights into the two dimensions of political ideology separately. However, these distinct threads of research have not yet been synthesised into a cohesive two-dimensional theory of political ideology. As well as explaining the evolutionary origins of the two dimensions, such a synthesis would ideally explain why this specific two-dimensional structure organises political attitudes and values, as opposed to some other structure, and would provide a rigorous evolutionary account of variation along the two dimensions within human populations. 

To organise, integrate, and expand on this previous work, we begin with the premise that politics is fundamentally the process of dealing with the conflicts of interest that arise from human group living [@Petersen2015; @Haidt2012; @Hibbing2013]. To understand the foundations of political ideology, one must therefore understand how human group living has evolved. Independent convergent evidence from anthropology, primatology, and developmental psychology suggests that, following the divergence from great apes, human group living evolved via two key shifts. First, humans began to cooperate more across wider interdependent networks and share the spoils of cooperation more evenly. Second, humans became more committed to group viability, conforming to social norms in culturally marked groups and punishing norm violators. These two social drives are strikingly concordant with two dimensions of political ideology captured in Table \@ref(tab:ch2tabScales).

## Two key shifts in the evolution of human group living

The socio-political lives of great apes are complex [@DeWaal1982]. Chimpanzee social groups, for example, are organised by dominance hierarchies. Owing to the fitness benefits of higher status within these hierarchies, rank positions are hotly contested and change dynamically over time [@Foerster2016], with individuals frequently engaging in Machiavellian social strategies to contest the status quo [@DeWaal1982]. Chimpanzees also patrol territorial borders to defend their group against outsiders [@Watts2001]. Much like humans, then, the political lives of great apes are spent dealing with the challenges of group living.

Human group living shares much of this complexity. However, the ancestral human hunter-gatherer communities that emerged over the course of the Pleistocene were vastly different from those of other great apes. They were characterised by contact and trade between extended networks [@Bird2019; @Hill2011; @Lee1972], relatively egalitarian socio-political structures [@Boehm1993], and deeply embedded cultural norms, conventions, and institutions [@Henrich2015]. To explain the emergence of this unique social organisation, researchers have posited two key shifts in the evolution of human group living [@Jensen2014; @Sterelny2007; @Sterelny2016; @Tomasello2012; @Tomasello2013; @Tomasello2016] that are thought to have paved the way for more complex human societies.

In the first key shift, humans began to cooperate more across wider interdependent networks and share the spoils of cooperation more evenly. Resource scarcity in the Pleistocene required early humans to forage collaboratively for rarer but higher-value calorie-dense foods [@Tomasello2012]. As fitness interdependence [@Aktipis2018] with group members increased, selection favoured those who could readily work together with and provide benefits for others. Shifting to a riskier mode of subsistence also selected for egalitarian meat-sharing as a form of risk pooling [@Gurven2004], replacing the great ape system of resource distribution based on dominance and hierarchy. These changes resulted in a human psychology uniquely sensitive to cooperative interactions with others. Building upon psychological mechanisms for kin altruism, reciprocity, and reputation management, humans developed other-regarding preferences and empathic concern [@Jensen2014] that allowed them to extend cooperation beyond kith and kin to a wider network of interdependent individuals. In line with this, evidence suggests that, unlike other great apes [@Krupenye2018; @Silk2005; @Warneken2006a], humans spontaneously help others [@Warneken2006b], effectively communicate to solve coordination problems [@Cooper1992; @Koomen2018], prefer egalitarian divisions of resources [@Gurven2004; @Hamann2011], and favour cooperative over exploitative individuals [@Hamlin2007].

In the second key shift, humans became more group-minded, conforming to social norms in culturally marked groups and punishing norm violators. Throughout the Pleistocene, human groups began to expand in size and compete with rival groups, further increasing fitness interdependence within groups [@Tomasello2012]. To deal with the coordination problems inherent to larger groups, social norms created the conventions, common knowledge of conventions, and shared meta-knowledge necessary for group-wide joint action [@Cronk2013]. In response to inter-group competition, individual-level selection favoured a sense of group identity, conformity to group norms, and punishment of norm violators to promote group cohesion in the face of external threats [@Mathew2011]. These changes resulted in a 'group-minded' human psychology underlain by self-conscious normative emotions, such as shame, group pride, and moral disgust [@Haidt2012; @Vaish2011], as well as psychological mechanisms for cultural learning within groups [@Henrich2015]. In line with this, evidence suggests that, unlike other great apes [@Burkart2018; @Riedl2012; @Tomasello2012], humans conform to group-wide social norms [@Claidiere2012], punish third-parties who violate social norms [@Fehr2004], and discern group membership by attending to cultural markers like religion [@Galen2011], language [@Kinzler2012], and accent [@Kinzler2011].

These two foundations of human group living are thought to have allowed us to transition from small kin bands of great apes to the larger, more socially complex communities of Pleistocene hunter-gatherers [@Tomasello2012]. But these communities were still small relative to modern human societies. Over the course of the Holocene, a process of cultural evolution following the advent of agriculture transitioned the default mode of group living from hunter-gatherer communities to ethnically and economically diverse mega-cities and nation states [@Sterelny2016]. Yet beneath these recent cultural innovations, the cognitive mechanisms underlying human cooperation and group conformity have remained largely unchanged.

These two fundamental human responses to the challenges of group living --- cooperation and group conformity --- explain why scholars have repeatedly identified a two-dimensional structure of political ideology and provide a principled foundation for the domain of each dimension. The first domain (left column of Table \@ref(tab:ch2tabScales)) is concerned with cooperating more across wider interdependent networks and sharing the spoils of cooperation more evenly. In our ancestral past, individuals had to constantly navigate cooperative dilemmas, such as collaborative foraging and meat sharing, as well as determine how to share the spoils of cooperation. Today, analogous dilemmas underlie policy issues like taxation, welfare programs, and free education. The second domain (right column of Table \@ref(tab:ch2tabScales)) is concerned with group conformity. For early humans living in highly interdependent social groups, it was vital to abide by group-wide social norms, sanction norm violators, and defend the group against outsiders. Today, analogous concerns about group viability will manifest themselves in attitudes regarding traditional social values, criminal justice, patriotism, and national security. Hence, cooperation and group conformity provide the dual evolutionary foundations of modern political ideology.

## Variation in cooperative and group conformist behaviour

As well as explaining how cooperation and group conformity came to be species-typical human social drives, an evolutionary approach provides a natural framework and set of mechanisms for understanding observed variation in such traits [@Buss2009]. Here we consider two mechanisms --- fitness trade-offs and behavioural plasticity --- to show how both genes and environment together predict strategic variation in cooperation and group conformity in human populations.

Fitness trade-offs exist when extreme levels of a trait confer both benefits and costs to individuals. Such trade-offs can lead to the evolution of functional, heritable individual differences via fluctuating selection [@Nettle2006]. In other words, variation in a trait is preserved if different levels of the trait provide different benefits in different contexts. For example, researchers have attributed personality variation in both humans [@Nettle2006] and non-human animals [@Wolf2007] to fitness trade-offs. In humans, high levels of extraversion are associated with a greater number of sexual partners, but also with greater risk of accident or illness [@Nettle2005]. This trade-off results in an extraversion spectrum along which individuals can vary.

In a similar vein, fitness trade-offs have shaped individual differences in the basic social drives for cooperation and group conformity in human populations. Model-based simulations of evolution show that uncertain social environments can select for general drives ('heuristics') for cooperation and group conformity [@Bear2016; @Richerson2005; @Delton2011; @Gigerenzer2008; @Krasnow2013; @VandenBerg2018]. Researchers argue that these general drives apply across different situational contexts and are proximately motivated by other-regarding preferences and empathic concern (cooperation) or shame, group pride, and moral disgust [group conformity; @Haidt2012; @Jensen2014; @Tomasello2013]. As with extraversion, the general drives for cooperation and group conformity come with both costs and benefits. Individuals with a general cooperative drive benefit from good reputations and subsequent partner choice, but leave themselves open to exploitation from free-riders. Individuals with a general conformist drive benefit from adaptability to the group's local conditions and increased group viability, but sacrifice possibilities for individual learning and innovation [@Rogers1988]. These fitness trade-offs are expected to maintain heritable individual differences in general drives for both cooperation and group conformity within human populations. In line with this, individual differences in cooperation are heritable [@Cesarini2008], and cooperative behaviour is positively correlated across a variety of anonymous one-shot economic games [@Carlsson2014; @McAuliffe2019; @Peysakhovich2014; @Yamagishi2013] and predicts self-reported trust and altruism [@Johansson-Stenman2013; @McAuliffe2019], as well as a range of real-world cooperative behaviours [@Benz2008; @Rustagi2010]. Similarly, individual differences in norm-enforcing punishment are heritable [@Wallace2007], and conformist behaviour is positively correlated across various measures of norm enforcement and norm adherence [@Kimbrough2016; @Peysakhovich2014].

In addition to heritable individual differences, phenotypic variation in willingness to cooperate and conform is also expected as an adaptive response to the social environment, either on-the-fly or canalized in early development, a phenomenon usually referred to as behavioural plasticity [@Snell-Rood2013]. There is reason to believe that both humans and non-human animals calibrate their cooperative and conformist behaviour based on feedback from their social environment. With regard to cooperation, psychological mechanisms integrate cues about both the situation (e.g., the presence of an audience) and the target of the interaction (e.g., whether the target previously defected) when determining whether to cooperate [@Delton2016]. For example, both chimpanzees [@Cronin2014] and humans [@Cronin2015] are less cooperative in hierarchical or competitive social environments, likely because they perceive potential targets as less trustworthy partners for long-term cooperative relationships. With regard to group conformity, psychological mechanisms integrate cues about environmental unpredictability and threat when determining whether to conform to the group. Nine-spined sticklebacks [@VanBergen2004], rats [@GalefJr2008], and humans [@Morgan2011] all engage in conformist social learning when environments are perceived to be unpredictable with a high cost of individual learning. Furthermore, in humans, several classes of threat have been shown to increase group conformity. Cues of disease-causing pathogens motivate conformity [@Murray2012; @Wu2012] to reduce the risk of infection [@Henrich2015]. Cues of dangerous situations motivate group conformity as self-protection [@Griskevicius2006]. Cues of threats to group viability (e.g., inter-group conflict, rapid immigration) motivate conformity in the form of adherence to religious norms [@Henrich2019], punishment of in-group norm violators [@Gneezy2012], and out-group prejudice [@Stenner2005], maintaining group solidarity and increasing in-group coordination efficiency. Thus, behavioural plasticity is expected to adaptively calibrate individuals' willingness to cooperate and conform.

We propose that both heritable individual differences and functional behavioural plasticity have maintained strategic variation in cooperation and group conformity in human populations, naturally giving rise to variation along the two dimensions of political ideology. Heritable individual differences in a general cooperative drive, combined with the presence of relevant situational cues (e.g., perceptions of targets' need or ability to reciprocate) produce variation in willingness to cooperate with others, giving rise to variation along the first dimension of political ideology (left column of Table \@ref(tab:ch2tabScales)). Heritable individual differences in a general group conformist drive, combined with the presence of relevant situational cues (e.g., perceived unpredictability or threats to group norms), produce variation in willingness to conform to and enforce group-wide social norms and defend the group, giving rise to variation along the second dimension of political ideology (right column of Table \@ref(tab:ch2tabScales)). Hence, in addition to explaining the essential nature of the two dimensions of political ideology, an evolutionary approach outlines how predictable interactions between genes and environment produce variation along the two dimensions.

## Cooperation, group conformity, and political ideology

The dual evolutionary framework illuminates existing work in political psychology, explaining why scholars have repeatedly converged on two dimensions of ideology, one referring to cooperation and the other referring to group conformity (Table \@ref(tab:ch2tabScales)). A close look at some exemplar items from the self-report scales in Table \@ref(tab:ch2tabScales) reveals this pattern more clearly (Table \@ref(tab:ch2tabScaleItems)). The scale items in the left column of Table \@ref(tab:ch2tabScaleItems) measure willingness to cooperate with others. Some items refer to helping, empathy, and prosocial concern, qualities that would have been crucial prerequisites for collaboration in early human groups. Other items emphasise egalitarianism, equality, and fairness, reminiscent of the kinds of anti-hierarchical levelling mechanisms early humans would have employed when sharing the spoils of cooperation [@Tomasello2012]. Solipsism and unity with nature represent constrained versus extended cooperative networks, respectively. The reverse-coded scales (social dominance orientation and the beta-isms) describe competitive tendencies, such as self-interested and dominating behaviour, that aim to distribute resources according to hierarchy. The scale items in the right column of Table \@ref(tab:ch2tabScaleItems), by contrast, measure adherence to group-wide social norms, punishment of in-group norm violators, and parochialism. Many items focus on traditionalism, obedience, and deference to authority, which can be understood as outcomes of psychological predispositions for majority-biased and prestige-biased conformist learning within groups [@Henrich2004]. Several items emphasise strict laws, justice, and penalties for offenders, which clearly relate to norm enforcement. Other items refer to patriotism and the need for national security, reflecting parochial in-group favouritism and concern for group viability.

(ref:ch2ref22) __Economic conservatism (core issues) [@Feldman2014]__

(ref:ch2ref23) __Social conservatism (core issues) [@Feldman2014]__

(ref:ch2ref24) __Social dominance orientation [@Pratto1994]__

(ref:ch2ref25) __Right-wing authoritarianism [@Altemeyer1981]__

(ref:ch2ref26) __Unmitigated self-interest ('beta-isms') [@Saucier2000]__

(ref:ch2ref27) __Tradition-oriented religiousness ('alpha-isms') [@Saucier2000]__

(ref:ch2ref28) __Self-enhancement vs. self-transcendence [@Schwartz1996]__

(ref:ch2ref29) __Conservation vs. openness [@Schwartz1996]__

(ref:ch2ref30) __Individualising (care-harm, fairness-reciprocity) [@Graham2009]__

(ref:ch2ref31) __Binding (in-group-loyalty, authority-respect, sanctity-purity) [@Graham2009]__

(ref:ch2ref32) @Claessens2020a

```{r ch2tabScaleItems, echo=F}
knitr::kable(
    readd(ch2_tabScaleItems), longtable = TRUE, booktabs = TRUE,
    format = "latex",
    caption = "\\textit{Item exemplars from a subset of scales measuring the two dimensions of political ideology.} Table adapted from (ref:ch2ref32).",
    col.names = gsub("[.]", " ", names(readd(ch2_tabScaleItems))),
    escape = FALSE
  ) %>%
  kable_styling(font_size = 10) %>%
  column_spec(1:2, width = "18em")
```

The dual evolutionary framework also makes sense of the policy stances that these ideological scales predict. Economic conservatism predicts stances on issues like government-funded healthcare, welfare programs, and free education [@Feldman2014]. Though far removed from the cooperation problems faced by early human groups, these issues can all be framed as cooperative dilemmas, in which an individual's short-term self-interest is at odds with the group's long-term collective interest [@VanLange2013]. For such issues, people's willingness to cooperate versus compete with others influences their political views. For example, individuals with a greater willingness to cooperate are more likely to support extra taxes to fund a healthcare system that is accessible to everyone in the group (a cooperative dilemma). Social conservatism, by contrast, predicts stances on issues like traditional social values, criminal justice, national security, and religion [@Altemeyer1981; @Feldman2014]. Group conformity underlies all these political stances. Social conservatives are more likely to conform to their group's traditional social norms (e.g., family structures, gender roles, and marriage norms), support policies that increase the influence of these norms in the public sphere [@Altemeyer1981], and endorse punitive rather than rehabilitative action towards criminals and other in-group norm violators [@Capps2002; @Feather1996]. They often support tougher border control and military intervention abroad, as they are keenly aware of cultural group boundaries and are motivated to maintain a viable in-group in the presence of perceived out-group threats. Norm adherence and norm enforcement in social conservatives is also often tied up with religion [@Altemeyer1981]. Anthropologists have long recognised religion as partly functioning to enforce sacred group norms and thus create moral communities [@Durkheim1912]. Similarly, evolutionary theorists have argued that religions are culturally group-selected packages of norms outlining which behaviours are permissible and how norm violators should be punished [@Norenzayan2016].

Even for more complex issues that do not relate straightforwardly to one dimension or the other, the dual evolutionary foundations of political ideology can help us unravel the psychological motivations behind the patterns we observe. Anti-immigration views, for example, can be associated with economic and/or social conservatism [@Duckitt2009]. The dual evolutionary framework predicts that these associations reflect different motives, with economic conservatives more sensitive to the possibility that successful immigrants will compete with them for resources and threaten the existing hierarchy, and social conservatives more concerned with the potential for cultural deviance and incompatibility of social norms, threatening group viability. Consistent with these predictions, social (but not economic) conservatives from the US and Switzerland were found to have less favourable attitudes towards immigrants who they thought were unlikely to assimilate to their group's norms, whilst economic (but not social) conservatives had less favourable attitudes towards immigrants who they thought were more likely to assimilate and thereby compete for resources and status [@Thomsen2008]. When immigrants are seen as a threat to egalitarian safety nets like the welfare system, economic progressives should also oppose immigration; in fact, precisely this justification was offered when Denmark's Social Democratic party announced policies seen as anti-immigration [@Henley2019].

With its emphasis on social norms, the dual evolutionary framework acknowledges that the attitudes of social conservatives should differ depending on the particular norms present in their society. However, this account does not reduce to cultural constructivism. Social norms are not entirely arbitrary; they often govern fitness-relevant behaviours (e.g., pathogen avoidance, mate choice, and reproduction) and group viability. For example, Fijian food norms that forbid pregnant women from ingesting toxic marine species are cultural adaptations that avoid deadly foetal poisoning [@Henrich2015]. Likewise, religious norms surrounding infidelity, abortion, and sexual orientation can be seen as culturally evolved mechanisms that promote larger families and increase group size [@Norenzayan2016]. Thus, this framework explains why social conservatives in the United States can become focused on behaviours such as marriage, contraception, prayer in school, and alcohol and drug use [@Altemeyer1981; @Feldman2014]: they are adhering to and enforcing social norms that govern fitness-relevant behaviours and group viability.

Cooperation is itself a fitness-relevant behaviour that social norms can govern. As a result, the dual evolutionary framework predicts that the phenotypic expression of economic conservatism (cooperation) and social conservatism (group conformity) may be correlated within populations, but the strength and direction of this relationship will vary across populations (i.e., the two dimensions are distinct). Consistent with these predictions, in developed Western democracies, economic and social conservatism are often weakly-to-moderately positively correlated with one another among political elites. For example, the Republican Party in the United States supports both economically conservative policies (e.g., free-market capitalism and less taxation) and socially conservative policies (e.g., national defence and opposition to gay marriage). However, in many cultures around the world, economic and social conservatism are weakly negatively correlated [@Malka2019]. For example, the Fidesz party in Hungary supports economically progressive policies (e.g., minimum wage increases and equitable pension systems) but socially conservative policies [e.g., strong border control; @Bayer2018]. This negative correlation is stronger in post-communist countries, like Hungary, where traditional social norms promote national equality, egalitarianism, and fairness [@Malka2019]. Intriguingly, at the individual level, the negative correlation is stronger among politically unengaged people across countries [@Johnston2018; @Malka2019], suggesting this may be the default organisation of the two dimensions in the absence of clear norms or cues from political elites. This is expected under the dual evolutionary framework if people readily interpret more egalitarian norms as in the interests of the group.

An understanding of the fitness trade-offs associated with the evolution of cooperation and group conformity makes sense of individual differences in economic decision-making, personality traits, and neurophysiology, and it explains why this variation reliably correlates with political ideology. Stable individual differences in both cooperation and norm-enforcing punishment in experimental economic games have been linked to political ideology [e.g., social dominance orientation; @Haesevoets2015; @Halali2018; @Thielmann2020] and real-world social values regarding taxation and revenge [@Peysakhovich2014]. Economic conservatism covaries with personality traits like agreeableness [@Heaven2001], social value orientation [@Balliet2016], and Machiavellianism [@Jones2013], while social conservatism covaries with openness to experience [@Duckitt2010b] and need for closure [@Jost2003]. Individual differences in basic neural and physiological processes also covary with ideology. When viewing images of others in distress, people higher in economic conservatism show less activation in brain regions associated with empathic concern [@Chiao2009]. People higher in social conservatism are also more likely to attend to and respond to threatening stimuli [@Dodd2012; @Hibbing2014; @Oxley2008]. The dual evolutionary framework explains these individual differences and correlations with political ideology as resulting from fitness trade-offs in general drives for cooperation and group conformity.

Similarly, the dual evolutionary framework provides insight into the influence of socio-environmental context on political ideology [@Duckitt2009]. Psychological mechanisms designed to integrate cues of potential long-term cooperative partnerships [@Delton2016] should cause people to adapt their levels of cooperation based on the amount of competition they perceive in their environment. Consistent with this prediction, individuals who view the world as a 'competitive jungle' score higher on measures of economic (but not social) conservatism [@Duckitt2009]. Similarly, people should adapt their levels of conformity based on perceived environmental unpredictability and cues of infectious diseases, dangerous situations, and threats to group viability. Consistent with this prediction, those who view the world as threatening, dangerous, and unpredictable score higher on measures of social (but not economic) conservatism [@Duckitt2009].

Finally, the dual evolutionary framework provides a theoretical scaffold for prior evolutionary approaches to the two dimensions of political ideology. Economic conservatism correlates with upper-body strength and resource-holding power [@Petersen2019] because these are important cues for the psychological mechanisms underlying cooperation. If an individual correctly perceives their own resource-holding power as high, this will motivate the accumulation of resources via dominance and power rather than egalitarian sharing [@Geniole2017], thus predicting economic (but not social) conservatism. By contrast, social conservatism correlates with disgust sensitivity and threat sensitivity [@Hibbing2014; @Tybur2016] because disease-causing pathogens and threats to individual safety and group viability are important cues for the psychological mechanisms underlying group conformity.

## Discussion

The dual evolutionary framework of political ideology proposes that two key human adaptations to group living --- cooperation and group conformity --- are the evolutionary foundations underlying the two repeatedly identified dimensions of political ideology in humans. It is possible that these two dimensions alone are not sufficient to capture the full breadth of political views. Several evolutionary approaches have posited three [@Sinn2017; @Tuschman2013], five [@Haidt2012], or even six [@Iyer2012] underlying dimensions. However, more complex factor models are not supported by the data [@Curry2019] and are readily reduced down to two dimensions when using principal components analyses and other factor-analytic statistical methods [@Federico2013; @Glover2014; @Sinn2017]. Other promising approaches in political psychology have attempted to carve the two dimensions into distinct sub-dimensions [@Duckitt2010a; @Ho2015]. For example, right-wing authoritarianism has been split into authoritarian submission, conventionalism, and authoritarian aggression [@Duckitt2010a]. Consistent with the dual evolutionary framework, these can be understood as an evolved commitment to group viability via conformity to existing group norms, conformity to traditional group norms, and punishment of norm violators, respectively [@Kessler2008]. While such approaches add nuance, the strong and reliable positive correlations between these sub-dimensions [@Duckitt2010a; @Ho2015] suggest that they represent two coherent packages of social motives that act together to organise cooperative and conformist behaviour.

Some scholars have suggested that, since most people in the electorate are unable to articulate why they harbour particular attitudes, the public is largely non-ideological [@Converse1964; @Zaller1992]. While it is true that not everyone is politically knowledgeable, aware, and engaged, a lack of political sophistication in the population should not be used as evidence against individual variation showing an underlying structure [@Jost2006]. Much like the use of language without the metacognitive awareness of its grammatical rules, people can hold ideologically consistent political attitudes without any explicit awareness of their structure.

Other scholars have argued that the structure of political attitudes emerges not from an underlying ideology but from the interplay between (inclusive) fitness interests and coalitions constructed by political elites [@Petersen2015; @Weeden2014]. Under this account, people simply align with political parties that best advance their fitness interests and then form political coalitions around clusters of interests. While self-interest undoubtedly plays a role in policy preferences (e.g., support for lower tuition fees among students versus support for retirement benefits among the elderly) and political elites clearly drive some aspects of policy organisation (e.g., the Republican and Democratic parties aligning economic and social policies), this account is, at best, incomplete. The two dimensions of political ideology are heritable and emerge across cultures [@Ashton2005; @Batricevic2017; @Lewis2014], suggesting an underlying structure that is not arbitrarily packaged by political elites. Furthermore, policy views frequently contradict people's fitness interests. For example, social dominance orientation is only marginally related to socio-economic status [@Pratto1994], and wealthy individuals often support economic redistribution while disadvantaged individuals often oppose welfare policies [@Jost2003].

Nevertheless, people do naturally group together with others who share their political views and show intense disliking for groups with inconsistent views [@Brandt2014]. Although there is some evidence that social conservatives are more sensitive to the 'value violations' motivating this ideological conflict [@Czarnek2019], even economic and social progressives reveal their intolerance if asked their opinions on wealthy business people, pro-lifers, and religious fundamentalists. The dual evolutionary framework suggests that this dislike stems from the fundamental conflicts between cooperation versus competition and individualism versus conformity as opposing strategies for navigating human group living. In line with this, recent work has shown that political intolerance is "dimension-specific" [@Crawford2017b]. Economic progressives and economic conservatives are intolerant of one another's views on cooperative dilemma issues. Separately, social progressives and social conservatives are intolerant of one another's views on normative issues.

## Conclusion

To conclude, we hope to encourage a fruitful dialogue between evolutionary scholars and political scientists to further our understanding of the foundations of political ideology. Political scientists have made great complementary strides in studying the two-dimensional structure of ideology, but questions have remained about why this particular structure exists. Evolutionary theory provides the meta-theoretical tools to answer such questions. The framework presented here shines light on existing work in political psychology and offers insight into the volatility in our current political climate. To return to a previous example, many of the within-party disagreements in the 2016 US presidential election can be understood as outcomes of the two dimensions of ideology. Hillary Clinton and Bernie Sanders diverged on egalitarian issues like taxation, healthcare, and free higher education. Before running for office, Donald Trump differed from other Republicans in his less-than-stringent approach to traditional social norms regarding abortion and same-sex marriage. We are optimistic that, in both political science and everyday public discourse, the multidimensional evolutionary framework we have presented will provide a more nuanced understanding of the politics that both unites and divides us.
