# Cooperative and conformist behavioural preferences predict the dual dimensions of political ideology^[Chapter based on "Claessens, S., Sibley, C. G., Chaudhuri, A., & Atkinson, Q. D. (under review). Cooperative and conformist behavioural preferences predict the dual dimensions of political ideology."]

```{r setup, include = FALSE}
loadd(ch3_d1)
loadd(ch3_d2)
```

## Introduction

Humans differ profoundly in their views on political issues like income redistribution, taxation, welfare, abortion, and religious instruction in schools. With an increasing focus on political polarisation and the consequent challenges of finding common ground on a host of pressing issues [@Neal2018; @PewResearchCenter2014], understanding the sources of this variation is more pressing than ever. One prominent explanation for variation in political views is that liberals and conservatives differ fundamentally in psychological traits and social preferences [@Hibbing2014; @Jost2003]. For example, research suggests that, compared to liberals, conservatives exhibit increased sensitivity to threats [@Hibbing2014], personal needs for closure [@Jost2003], and intolerance of uncertainty [@Jost2017], producing variation in support for policies that promote order, structure, and stability. By contrast, other evidence suggests that liberals and conservatives do not differ psychologically [@Bakker2020; @Brandt2014; @Crawford2017a; @Crawford2017b] and that variation in political views between liberals and conservatives instead reflects expressions of self-interest based on socio-demographic factors, identity, and political coalitions [@Weeden2014].

The debate about the existence of liberal-conservative psychological differences has largely treated political ideology as a unidimensional spectrum from left to right [but see @Crawford2017b]. But decades of interdisciplinary work indicate that political ideology is best described along two dimensions [@Ashton2005; @Claessens2020a; @Duckitt2009; @Feldman2014; @Saucier2000; @Schwartz1992]. The first dimension captures views on issues like taxation, welfare, capitalism, and big business [@Feldman2014], and has been variously labelled economic conservatism, social dominance orientation (SDO), individualising moral foundations, and egalitarianism [@Claessens2020a; @Duckitt2009; @Graham2009]. The second dimension captures views on issues like abortion, patriotism, sexual morality, and criminal justice [@Feldman2014], and has been labelled social conservatism, right-wing authoritarianism (RWA), binding moral foundations, and group loyalty [@Claessens2020a; @Duckitt2009; @Graham2009]. These two dimensions have repeatedly emerged from independent lines of research in political psychology [@Duckitt2009], moral psychology [@Graham2009], and cross-cultural psychology [@Ashton2005]. While the two dimensions of political ideology tend to positively covary in Western societies [@Duckitt2009], they are distinct [@Duckitt2009] and can vary independently, being negatively correlated with one another in many societies around the globe [@Malka2019].

Building on previous evolutionary approaches to politics and human sociality [@Haidt2012; @Hibbing2013; @Petersen2015], we recently proposed that the dual dimensions of political ideology can be understood as a product of two key trade-offs inherent to the evolution of human group living [@Claessens2020a]. This dual evolutionary framework argues that the first dimension of political ideology reflects a fundamental trade-off between cooperative egalitarianism and competitive hierarchy, and the second dimension reflects a fundamental trade-off between individualism and group conformity (i.e., reliance on socially-acquired information and adherence to social norms within cultural groups). These trade-offs are expected to produce phenotypic variation in general "heuristic" behavioural preferences for cooperation and group conformity that apply across different situational contexts. In line with approaches claiming that psychological differences exist across the political landscape, one prediction of the dual evolutionary framework is that variation in these general preferences for cooperation and group conformity will differentially predict the two dimensions of ideology in modern humans [@Claessens2020a]. Under this framework, an increased general preference for cooperation beyond close kin results in greater support for policies that promote large-scale cooperation and a more egalitarian distribution of resources, such as income redistribution, taxation, welfare, and pro-environmentalism. Likewise, an increased general preference for group conformity and viability results in greater support for policies that promote in-group conformity, norm enforcement, and the interests of the in-group, such as religious instruction in schools, capital punishment, and military spending.

If general preferences for cooperation and group conformity do give rise to differences in political opinion in modern humans, such differences should be manifest in preferences measured by incentivised behavioural tasks. Incentivised behavioural tasks (e.g., economic games) allow researchers to study variation in enduring social preferences [@Chaudhuri2009; @Pisor2020]. Decades of work using incentivised behavioural tasks have revealed individual differences in both cooperative preferences [@Kuhlman1975; @Kurzban2001; @Fischbacher2001; @Chaudhuri2011] and conformist preferences [@Molleman2019; @Toelch2014; @McElreath2005; @Efferson2008; @Molleman2014] in humans.

Previous research has studied the covariation between behaviour in incentivised tasks and the two dimensions of political ideology. Studies linking cooperation in economic games to unidimensional political party support have produced mixed results, with some studies finding that supporters of left-leaning parties are more cooperative in economic games [@Muller2019; @Kerschbamer2020; @Grunhage2020; @Cappelen2017] and other studies finding no relationship between gameplay and political party support [@Anderson2005; @Balliet2016]. However, a number of studies find a specific association between cooperative preferences and SDO, an ideological measure of preferences for hierarchy and anti-egalitarianism [@Haesevoets2015; @Halali2018]. In a meta-analysis of the links between personality and prosocial behaviour, @Thielmann2020 found that cooperative behaviour in a number of social dilemma games that allow exploitation of others was negatively correlated with SDO [@Duckitt2009]. In contrast, other research has failed to find a relationship between SDO and cooperative behaviour in the Prisoner's Dilemma Game, in which two players choose whether to cooperate or defect [@Balliet2016; @Haesevoets2015], or the Stag Hunt Game, in which two players must coordinate on one of multiple equilibrium strategies [@Haesevoets2015]. These findings suggest that SDO may relate to particular affordances of individual games [e.g., exploitation; @Thielmann2020] rather than to a general cooperative preference that applies across situations with different payoff structures, as predicted by the dual evolutionary framework.

Research exploring the relationship between political ideology and general conformist preferences is more limited. Several studies have found a relationship between conformist social learning strategies and collectivist values [@Toelch2014; @Molleman2019], and @Kemmelmeier2003 found that collectivist values are positively correlated with RWA, an ideological measure of authoritarian attitudes [@Duckitt2009]. In contrast, @Smirnov2010 found that norm-enforcing punishment in a collective action game did not differ between Republicans and Democrats, although these authors did not include measures of political ideology. Much like the work relating cooperative behaviour to political ideology, these studies are limited by only focusing on one or two incentivised behavioural tasks and single measures of political ideology.

Here, we extend prior work by combining a large suite of incentivised behavioural tasks with comprehensive survey data to test whether and to what extent cooperative and conformist behavioural preferences covary with measures of the two dimensions of political ideology and related policy preferences. Across two pre-registered studies, we recruited participants from the New Zealand Attitudes and Values Study, an ongoing nationally-representative longitudinal survey of the attitudes and values of New Zealanders [see Appendix to Chapter 3, Figure \@ref(fig:ch3samplePlot) for sample characteristics; @Sibley2020]. We asked a large diverse sample of participants (_N_ = 991) to complete a series of incentivised behavioural tasks measuring cooperative preferences, such as trust, altruism, and collective action, and conformist preferences, such as norm-enforcing punishment, social information use, and rule following. In Study 1, we used a battery of one-shot anonymous economic games that have previously been shown to capture two dimensions of behaviour, one reflecting a "cooperative phenotype" and another labelled "norm-enforcing punishment" [@Peysakhovich2014]. In Study 2, we extended this suite of games to include incentivised tasks capturing two further elements of conformist behaviour: social information use and rule following. Across these studies, we related cooperative and conformist preferences to SDO and RWA, a widely used pair of measures capturing the two dimensions of political ideology, respectively [@Duckitt2009]. We also related cooperative and conformist preferences to a host of economic and social policy views, respectively. By linking behaviour in these abstract tasks to a broad range of self-reported political attitudes, we aim to identify whether and to what extent cooperative and conformist preferences covary with the dual dimensions of political ideology.

```{r echo=F, warning=F, message=F}
loadd(ch3_cor)
loadd(ch3_pca1.2)
loadd(ch3_pca2.2)
```

## Study 1

### Methods

#### Participants

##### Power analysis

The minimum size of our sample was determined by conducting a power analysis on data from @Peysakhovich2014. We used the smallest significant correlation between economic game play and real-world cooperation in that study as the effect size (_r_ = 0.15). G*Power [@Faul2007] suggested a sample of 571 participants to detect this correlation effect size with statistical power of 0.95. We aimed to sample 1000 participants, substantially above this amount, and eventually report data from 991 participants.

##### Participants and sampling

Participants for this study were sampled from the ongoing longitudinal New Zealand Attitudes and Values Study [@Sibley2020]. This participant pool is a representative sample of the New Zealand population, contacted through random draws from the New Zealand electoral roll. From a sample frame of 3345 participants (see Appendix to Chapter 3, Supplementary Methods), we successfully contacted 2731 about the study. Participants were contacted initially by phone and then, if they agreed to participate, over email in the days leading up to their allocated study session. 1740 participants either dropped out of the study entirely (before or during the session), spent less than 5 minutes or more than 50 minutes on the games, or did not have SDO or RWA scores. This resulted in a final sample of `r nrow(ch3_d1)` participants (`r sum(ch3_d1$Gender.T10 == 0, na.rm = TRUE)` females; age _M_ = `r printnum(mean(ch3_d1$Age.T10, na.rm = TRUE), digits = 0)` years, _SD_ = `r printnum(sd(ch3_d1$Age.T10, na.rm = TRUE), digits = 0)` years; see Appendix to Chapter 3, Figure \@ref(fig:ch3samplePlot) for further sample characteristics).

#### Materials

##### New Zealand Attitudes and Values Survey measures

We took our main self-report variables from the 10^th^ wave of the New Zealand Attitudes and Values Study [@Sibley2020], an annual longitudinal survey which is completed by New Zealand participants both online and via post. The main self-report variables included SDO, RWA, age, gender, ethnicity, education level, socio-economic status, local deprivation, and religiosity. SDO and RWA scores were both mean scores of six separate Likert scale items (1 = strongly disagree, 7 = strongly agree). Additional items on policy views and competitive worldview were taken from earlier waves of the New Zealand Attitudes and Values Study. See Appendix to Chapter 3, Table \@ref(tab:ch3itemTable) for full list of self-report items.

##### Battery of economic games

Participants completed eight economic games, conducted online in real-time using oTree software [code available at https://osf.io/dwx8g/; @Chen2016]. In an attempt to replicate previous work, these games are largely identical to those from @Peysakhovich2014. All games involved one-shot decisions between multiple players. The strategy method was used to elicit responses in all possible roles. Participants played for points, which were converted to New Zealand dollars (1 point = $0.035).

We included three cooperation games in which individuals decided whether to incur a personal cost to benefit another player:

- _Dictator Game_. Player A is given 100 points. They must decide how many of these points to transfer to Player B. Player A keeps the remaining points. Player B is passive in the interaction.
- _Trust Game_. Players A and B both start with 50 points. First, Player A decides whether or not to transfer all 50 points to Player B, in the knowledge that the transferred amount will be tripled to 150 points. If Player A transfers, Player B now has 200 points. Player B must then decide to transfer between 0 and 150 points back to Player A.
- _Public Goods Game_. Four players begin with 100 points each. They can contribute between 0 and 100 points into a shared group project. All four decisions are made simultaneously, and then the amount in the group project is doubled and distributed evenly between all four players. Each player ends the game with their share from the group project, plus the points they initially refrained from contributing.

We also included three punishment games, in which individuals decide whether to incur a personal cost to punish another player for their decisions:

- _Ultimatum Game_. Player A starts with 100 points, and Player B starts with nothing. Player A must decide how many points to transfer to Player B. However, Player B simultaneously specifies their 'minimum acceptable offer': namely, the lowest transfer from Player A that they will accept. If Player A's transfer amount is lower than this minimum acceptable offer, both players end the game with 0 points. Otherwise, Player B receives the transfer amount, and Player A keeps the remaining points.
- _Third-Party Punishment Game_. Players A, B, and C all start with 100 points. First, Player A decides whether to 'take' from Player B. If Player A takes, Player B loses 50 points and Player A gains 30 points (taking is inefficient). If Player A takes, Player C can then pay between 0 and 20 points to remove points from Player A. Each paid point removes 5 points from Player A. Player B is passive in the interaction.
- _Second-Party Punishment Game_. Players A and B start with 100 points. This game has two stages: the transfer stage, and the penalty stage. In the transfer stage, each player decides whether to transfer 30 points to the other player. Any transferred points are doubled before the other player receives them. Decisions are made simultaneously. The transfer stage follows the payoff matrix of a Prisoner's Dilemma. Then, in the penalty phase, both players can pay between 0 and 10 points to remove points from the other player, depending on their decision in the transfer stage. Each paid point removes 5 points from the other player.

We replaced the destructive All-Pay Auction Game used in @Peysakhovich2014 with two coordination games: a version of the Stag Hunt Game, in which players must coordinate on one of multiple equilibria, with one of those equilibria dominating the others in terms of payoffs; and the Stag Hunt Game with Punishment, in which players may punish others for not choosing the strategy commensurate with the payoff-dominant equilibrium. These games are kept as identical to the Public Goods Game and Second-Party Punishment Game (respectively) as possible. We added these two games to study whether the general preferences for cooperation and punishment extended to coordination problems that do not involve exploitation of others and are arguably better models for the real-world cooperative dilemmas faced by our human ancestors [@Skyrms2003; @Tomasello2012]:

- _Stag Hunt Game_. Four players begin with 50 points each. They can either contribute 30 points into a shared group project or contribute nothing. Any points in the group project will be doubled and distributed evenly between all players, _but only if all players contribute_. Otherwise, the points in the group project will be lost. Decisions are made simultaneously. Each player ends the game with their share from the group project, plus the points they initially refrained from contributing.
- _Stag Hunt Game with Punishment_. Players A and B start with 100 points. This game has two stages: the transfer stage, and the penalty stage. In the transfer stage, each player decides whether to transfer 30 points to a group project. Any points in the group project will be doubled and distributed evenly between both players, _but only if both players contribute_. Otherwise, the points in the group project will be lost. Decisions are made simultaneously. The transfer stage follows the payoff matrix of a Stag Hunt. Then, in the penalty phase, both players can pay between 0 and 10 points to remove points from the other player, depending on their decision in the transfer stage. Each paid point removes 5 points from the other player.

#### Procedure

Data collection for Study 1 was conducted between 18^th^ February and 25^th^ July 2019 with weekly staggered recruitment. Study sessions contained between 14 and 97 participants, and were conducted on midweek evenings (see Appendix to Chapter 3, Supplementary Methods). Participants knew that they were playing with others recruited from the New Zealand Attitudes and Values Study, but were not aware of how many people were present in any particular session. Participants learned that they would be playing for points that would be later converted to currency, that they would be matched in real-time with other participants from the New Zealand Attitudes and Values Study, and that every task would count towards their final bonus payment.

The eight economic games were presented in a random order. For each game, participants read the instructions and answered a comprehension question about the game. Participants then provided responses for all possible roles in the game, following the strategy method. After all the games, participants entered a waiting lobby where they waited for other participants in the session to complete the games. Once all participants in the session arrived at the waiting lobby, the software calculated payoffs for each game by randomly matching individuals within a session. The accumulated payoffs across all eight games determined the final overall payoff.

Participants were paid a fixed \$20 New Zealand dollar show-up fee, plus a bonus payment of between \$10 and \$35 (_M_ = \$`r printnum(mean(ch3_d1$egame.Paid.T10 - 20))`, _SD_ = \$`r printnum(sd(ch3_d1$egame.Paid.T10 - 20))`) depending on the decisions made by them and their peers. On average, participants took `r printnum(mean(ch3_d1$egame.SecsAll.T10 / 60), digits = 0)` minutes to complete the eight economic games (_SD_ = `r printnum(sd(ch3_d1$egame.SecsAll.T10 / 60))` mins, range = `r printnum(min(ch3_d1$egame.SecsAll.T10 / 60), digits = 0)` - `r printnum(max(ch3_d1$egame.SecsAll.T10 / 60), digits = 0)` mins).

#### Pre-registration

We pre-registered three hypotheses on the Open Science Framework on 19^th^ November 2018 (https://osf.io/dwx8g/). First, we hypothesised that a two-factor structure of cooperation and punishment would emerge from the suite of economic games, replicating @Peysakhovich2014. Second, we hypothesised that SDO would negatively predict the cooperation factor. Third, we hypothesised that RWA would positively predict the punishment factor.

#### Statistical analysis

Our pre-registered analyses proceeded in two stages. In the first stage, we focused on the data from the economic games. Following @Peysakhovich2014, we limited our analyses to game decisions which were not followed by a subsequent punishment stage: so-called 'uncoerced' decisions. These decisions were scaled to vary between 0 and 1 for all analyses. We used a comprehensive set of statistical tests (correlations, principal components analyses, and confirmatory factor analyses) to determine the factor structure of variation in these decisions. We expected that two factors would emerge from and be supported by the data: cooperation and punishment. In the second stage, we used structural equation modelling to determine the extent to which these cooperation and punishment latent variables could be predicted by SDO and RWA. We also conducted additional exploratory analyses that were not pre-registered: (1) regressions predicting SDO and RWA from individual game decisions, and (2) structural equation models predicting policy views from the cooperation and punishment latent variables.

All analyses were conducted in R version 4.0.2 [@R-base]. We used the _psych_ package [@R-psych] for correlations and principal components analyses, and the _lavaan_ package [@R-lavaan] for confirmatory factor analyses and structural equation modelling. Figures were created with the _ggplot2_ [@R-ggplot2] and _cowplot_ [@R-cowplot] packages, and the reproducibility of all analyses was ensured by using the _drake_ package [@R-drake].

#### Ethical approval

Ethical approval for Study 1 was provided by the University of Auckland Human Participants Ethics Committee on 19^th^ July 2018 for three years (ref: 021666).

### Results

To test the first hypothesis that gameplay would produce a two-factor structure, we focused on patterns of covariation among the economic games with pre-registered correlational, principal components, and confirmatory factor analyses. Pairwise Spearman's rank correlations between game decisions, accounting for multiple comparisons, are visualised in Figure \@ref(fig:ch3corPcaPlot)a. Replicating @Peysakhovich2014, we found significant positive correlations between all cooperation decisions (_r_~s~ = `r printnum(min(ch3_cor$r[1:5,1:5][lower.tri(ch3_cor$r[1:5,1:5])]))`-`r printnum(max(ch3_cor$r[1:5,1:5][lower.tri(ch3_cor$r[1:5,1:5])]))`), significant positive correlations between all punishment decisions (_r_~s~ = `r printnum(min(ch3_cor$r[6:9,6:9][lower.tri(ch3_cor$r[6:9,6:9])]))`-`r printnum(max(ch3_cor$r[6:9,6:9][lower.tri(ch3_cor$r[6:9,6:9])]))`), and non-significant or negative correlations between most cooperation and punishment decisions. Principal components analyses suggested that these patterns of covariation could be explained by two underlying factors: cooperation and punishment. When including only games from previous work, principal components analysis supported a two-factor solution. Using orthogonal varimax rotation and retaining factors with eigenvalues above one (Appendix to Chapter 3, Figure \@ref(fig:ch3screePlot)a), this analysis revealed that cooperative behaviour loads highly onto one factor and punitive behaviour loads highly onto a second distinct factor (Figure \@ref(fig:ch3corPcaPlot)b). Together, these factors explained `r printnum(ch3_pca1.2$Vaccounted[3,2]*100, digits=0)`% of the variance in game decisions. This two-factor structure held when adding our novel coordination games, with cooperation in the Stag Hunt Game loading onto the cooperation factor and punishment in the Stag Hunt Game with Punishment loading onto the punishment factor (Figure \@ref(fig:ch3corPcaPlot)c and Appendix to Chapter 3, Figure \@ref(fig:ch3screePlot)b; `r printnum(ch3_pca2.2$Vaccounted[3,2]*100, digits = 0)`% variance explained).

(ref:ch3corPcaPlotCaption) _Factor structure of economic games._ (_a_) Pairwise Spearman's rank correlations for game decisions. Grey crosses indicate non-significant correlations with Benjamin-Hochberg corrected _p_-values above .05. (_b_) Factor loadings from a principal components analysis with only the game decisions from previous work [@Peysakhovich2014]. (_c_) Factor loadings from an extended principal components analysis including coordination games. DG = Dictator Game, TG1 = Trust Game (Give), TG2 = Trust Game (Return), PGG = Public Goods Game, SH = Stag Hunt Game, UG = Ultimatum Game (Minimum Acceptable Offer), TPP = Third-Party Punishment Game (Punish), SPP = Second-Party Punishment Game (Punish Defector), SHP = Stag Hunt Game with Punishment (Punish Defector).

```{r ch3corPcaPlot, echo=FALSE, warning=F, message=F, fig.width=8, fig.height=5, fig.cap="(ref:ch3corPcaPlotCaption)"}
readd(ch3_corPcaPlot)
```

```{r echo=F, warning=F, message=F}
loadd(ch3_cfa1)
loadd(ch3_cfa2)
```

In addition to the data-driven statistics reported above, the two-factor structure of cooperation and punishment was also supported by confirmatory factor analyses. As with our principal components analysis, we initially restricted our confirmatory factor model to include only the games used in previous work [@Peysakhovich2014]. Controlling for game comprehension, we loaded the Dictator Game, Trust Game (Give), Trust Game (Return), and Public Goods Game onto the cooperation latent variable, and loaded the Ultimatum Game (Minimum Acceptable Offer), Third-Party Punishment Game (Punish), and Second-Party Punishment Game (Punish Defector) onto the punishment latent variable. According to established fit statistic cutoffs [@Hu1999;@MacCallum1996], this model fitted the data well, with a Root Mean Square Error of Approximation (RMSEA) value of `r printnum(as.numeric(fitMeasures(ch3_cfa1)['rmsea']))`, a Standardised Root Mean Square Residual (SRMR) value of `r printnum(as.numeric(fitMeasures(ch3_cfa1)['srmr']))`, and a Comparative Fit Index (CFI) value of `r printnum(as.numeric(fitMeasures(ch3_cfa1)['cfi']))`. All indicators had significantly positive loadings, and the two latent variables were significantly negatively correlated with one another (_r_ = `r printnum(standardizedSolution(ch3_cfa1)[25,4])`, _p_ = `r printp(parameterEstimates(ch3_cfa1)[25,7])`; Appendix to Chapter 3, Figure \@ref(fig:ch3cfa1Plot)). Model fit improved further when additionally loading cooperation in the Stag Hunt Game onto the cooperation latent variable and loading punishment of defectors in the Stag Hunt Game with Punishment onto the punishment latent variable (RMSEA = `r printnum(as.numeric(fitMeasures(ch3_cfa2)['rmsea']))`; SRMR = `r printnum(as.numeric(fitMeasures(ch3_cfa2)['srmr']))`; CFI = `r printnum(as.numeric(fitMeasures(ch3_cfa2)['cfi']))`; Appendix to Chapter 3, Figure \@ref(fig:ch3cfa2Plot)).

```{r echo=F, warning=F, message=F}
loadd(ch3_sem1)
loadd(ch3_r_SDO_coop)
loadd(ch3_r_SDO_pun)
loadd(ch3_r_RWA_coop)
loadd(ch3_r_RWA_pun)
```

Having confirmed the existence of a two-factor structure underlying gameplay, we next fitted pre-registered structural equation models to test the relationships between the two latent variables and political ideology. We simultaneously regressed the cooperation and punishment latent variables onto mean scores of six Likert scale items for SDO (Cronbach's $\alpha$ = `r printnum(readd(ch3_alphaSDO))`) and RWA ($\alpha$ = `r printnum(readd(ch3_alphaRWA))`). In line with our second hypothesis, we found that SDO significantly negatively predicted the cooperation latent variable with a small-to-medium effect size (standardised $\beta$ = `r printnum(standardizedSolution(ch3_sem1)[19,4])`, unstandardised _b_ = `r printnum(parameterEstimates(ch3_sem1)[19,4])`, 95% confidence interval [`r printnum(parameterEstimates(ch3_sem1)[19,8])`, `r printnum(parameterEstimates(ch3_sem1)[19,9])`], _p_ `r printp(parameterEstimates(ch3_sem1)[19,7])`, semi-partial _r_ = `r printnum(ch3_r_SDO_coop)`; Figure \@ref(fig:ch3coopPunPlot)a). SDO also significantly positively predicted the punishment latent variable with a small effect size, which we did not hypothesise ($\beta$ = `r printnum(standardizedSolution(ch3_sem1)[21,4])`, _b_ = `r printnum(parameterEstimates(ch3_sem1)[21,4], digits=3)`, 95% CI [`r printnum(parameterEstimates(ch3_sem1)[21,8], digits=3)`, `r printnum(parameterEstimates(ch3_sem1)[21,9], digits=3)`], _p_ = `r printp(parameterEstimates(ch3_sem1)[21,7])`, semi-partial _r_ = `r printnum(ch3_r_SDO_pun)`). In contrast with our third hypothesis, we found no significant relationship between RWA and the punishment latent variable ($\beta$ = `r printnum(standardizedSolution(ch3_sem1)[22,4])`, _b_ = `r printnum(parameterEstimates(ch3_sem1)[22,4], digits=3)`, 95% CI [`r printnum(parameterEstimates(ch3_sem1)[22,8], digits=3)`, `r printnum(parameterEstimates(ch3_sem1)[22,9], digits=3)`], _p_ = `r printp(parameterEstimates(ch3_sem1)[22,7])`, semi-partial _r_ = `r printnum(ch3_r_RWA_pun)`; Figure \@ref(fig:ch3coopPunPlot)b). RWA was also unrelated to the cooperation latent variable ($\beta$ = `r printnum(standardizedSolution(ch3_sem1)[20,4])`, _b_ = `r printnum(parameterEstimates(ch3_sem1)[20,4], digits=2)`, 95% CI [`r printnum(parameterEstimates(ch3_sem1)[20,8], digits=2)`, `r printnum(parameterEstimates(ch3_sem1)[20,9], digits=2)`], _p_ = `r printp(parameterEstimates(ch3_sem1)[20,7])`, semi-partial _r_ = `r printnum(ch3_r_RWA_coop)`). This pattern of results held when additionally controlling for age, gender, ethnicity, education, socio-economic status, local deprivation, and religiosity (Appendix to Chapter 3, Table \@ref(tab:ch3semTable)).

(ref:ch3coopPunPlotCaption) _Relationships between model-predicted latent variable scores for cooperation and punishment and the two dimensions of political ideology._ (_a_) Social dominance orientation (mean score) is negatively related to the cooperation latent variable. (_b_) Right-wing authoritarianism (mean score) is unrelated to the punishment latent variable. Full lines are predictions from least-squares linear regressions without socio-demographic covariates and shaded areas are 95% confidence intervals.

```{r ch3coopPunPlot, echo=FALSE, warning=F, message=F, fig.width=8, fig.height=4, fig.cap="(ref:ch3coopPunPlotCaption)"}
readd(ch3_coopPunPlot)
```

```{r echo=F}
loadd(ch3_r_indGamesSDO)
rs <-
  c(
    ch3_r_indGamesSDO[["tpp1"]], ch3_r_indGamesSDO[["pgg"]],
    ch3_r_indGamesSDO[["sh"]],   ch3_r_indGamesSDO[["tg2"]],
    ch3_r_indGamesSDO[["dg"]],   ch3_r_indGamesSDO[["ug1"]]
  )
```

To further investigate the relationships between political ideology and gameplay, we conducted a series of exploratory linear regressions with mean SDO and RWA scores as outcome variables and individual game decisions as predictors, accounting for multiple comparisons and statistically controlling for socio-demographics and game comprehension. We found that most cooperative decisions were negatively associated with SDO (Appendix to Chapter 3, Figure \@ref(fig:ch3indGamesPlotSDO)). Holding all other variables constant, individuals with higher SDO scores gave less in the Dictator Game, returned less in the Trust Game, contributed less in the Public Goods Game, did not coordinate in the Stag Hunt Game, offered less in the Ultimatum Game, and were more likely to take in the Third-Party Punishment Game. The effect sizes for these relationships were small, but consistent across games (semi-partial _r_ = `r round(min(rs), 2)`-`r round(max(rs), 2)`). Contrary to our predictions, SDO was also positively related to punishment decisions in the Second-Party Punishment Game and Stag Hunt Game with Punishment. Fewer individual game decisions predicted RWA (Appendix to Chapter 3, Figure \@ref(fig:ch3indGamesPlotRWA)). Holding all other variables constant, individuals with higher RWA scores were more likely to take in the Third-Party Punishment Game and engaged in greater anti-social punishment in the Second-Party Punishment Game and the Stag Hunt Game with Punishment, where "anti-social punishment" is defined as costly punishment of cooperators.

```{r echo=F, warning=F, message=F}
loadd(ch3_policy01)
loadd(ch3_policy02)
loadd(ch3_policy03)
loadd(ch3_policy04)
loadd(ch3_policy05)
loadd(ch3_policy06)
loadd(ch3_policy07)
loadd(ch3_policy08)
loadd(ch3_policy09)
loadd(ch3_r_policy01_coop)
loadd(ch3_r_policy02_coop)
loadd(ch3_r_policy03_coop)
loadd(ch3_r_policy07_coop)
loadd(ch3_r_policy08_coop)
loadd(ch3_r_policy09_coop)
loadd(ch3_r_policy04_pun)
loadd(ch3_r_policy05_pun)
loadd(ch3_r_policy06_pun)
```

The cooperation latent variable also reliably predicted economically progressive views across a wide range of political issues (Figure \@ref(fig:ch3policyPlotCoop); Appendix to Chapter 3, Tables \@ref(tab:ch3policy01table) - \@ref(tab:ch3policy09table)). Controlling for the punishment latent variable, game comprehension, and socio-demographics, exploratory structural equation modelling revealed that the cooperation latent variable positively predicted preferences for income redistribution (_b_ = `r printnum(parameterEstimates(ch3_policy01)[19,4], digits=2)`, 95% CI [`r printnum(parameterEstimates(ch3_policy01)[19,8], digits=2)`, `r printnum(parameterEstimates(ch3_policy01)[19,9], digits=2)`], _p_ `r printp(parameterEstimates(ch3_policy01)[19,7])`, semi-partial _r_ = `r printnum(ch3_r_policy01_coop, digits=2)`), making sacrifices for the environment (_b_ = `r printnum(parameterEstimates(ch3_policy03)[19,4], digits=2)`, 95% CI [`r printnum(parameterEstimates(ch3_policy03)[19,8], digits=2)`, `r printnum(parameterEstimates(ch3_policy03)[19,9], digits=2)`], _p_ = `r printp(parameterEstimates(ch3_policy03)[19,7])`, semi-partial _r_ = `r printnum(ch3_r_policy03_coop, digits=2)`), increased payments to those receiving Jobseeker Support (_b_ = `r printnum(parameterEstimates(ch3_policy07)[19,4], digits=2)`, 95% CI [`r printnum(parameterEstimates(ch3_policy07)[19,8], digits=2)`, `r printnum(parameterEstimates(ch3_policy07)[19,9], digits=2)`], _p_ `r printp(parameterEstimates(ch3_policy07)[19,7])`, semi-partial _r_ = `r printnum(ch3_r_policy07_coop, digits=2)`), and increased payments to those receiving Sole Parent Support (_b_ = `r printnum(parameterEstimates(ch3_policy08)[19,4], digits=2)`, 95% CI [`r printnum(parameterEstimates(ch3_policy08)[19,8], digits=2)`, `r printnum(parameterEstimates(ch3_policy08)[19,9], digits=2)`], _p_ `r printp(parameterEstimates(ch3_policy08)[19,7])`, semi-partial _r_ = `r printnum(ch3_r_policy08_coop, digits=2)`). Cooperation was also negatively related to support for an economically conservative "flat tax" where everyone pays the same percentage of tax on their income regardless of their wealth (_b_ = `r printnum(parameterEstimates(ch3_policy09)[19,4], digits=2)`, 95% CI [`r printnum(parameterEstimates(ch3_policy09)[19,8], digits=2)`, `r printnum(parameterEstimates(ch3_policy09)[19,9], digits=2)`], _p_ = `r printp(parameterEstimates(ch3_policy09)[19,7])`, semi-partial _r_ = `r printnum(ch3_r_policy09_coop, digits=2)`) and the belief that people would be less motivated to work hard if incomes were equal (_b_ = `r printnum(parameterEstimates(ch3_policy02)[19,4], digits=2)`, 95% CI [`r printnum(parameterEstimates(ch3_policy02)[19,8], digits=2)`, `r printnum(parameterEstimates(ch3_policy02)[19,9], digits=2)`], _p_ `r printp(parameterEstimates(ch3_policy02)[19,7])`, semi-partial _r_ = `r printnum(ch3_r_policy02_coop, digits=2)`). Though these effect sizes are small, the cooperation latent variable explains a comparable proportion of variance to other socio-demographic variables (Appendix to Chapter 3, Figure \@ref(fig:ch3policyR2Plot)). In contrast, the punishment latent variable did not predict socially conservative views like support for same-sex marriage (_b_ = `r printnum(parameterEstimates(ch3_policy04)[20,4], digits=2)`, 95% CI [`r printnum(parameterEstimates(ch3_policy04)[20,8], digits=2)`, `r printnum(parameterEstimates(ch3_policy04)[20,9], digits=2)`], _p_ = `r printp(parameterEstimates(ch3_policy04)[20,7])`, semi-partial _r_ = `r printnum(ch3_r_policy04_pun, digits=2)`), euthanasia (_b_ = `r printnum(parameterEstimates(ch3_policy05)[20,4], digits=2)`, 95% CI [`r printnum(parameterEstimates(ch3_policy05)[20,8], digits=2)`, `r printnum(parameterEstimates(ch3_policy05)[20,9], digits=2)`], _p_ = `r printp(parameterEstimates(ch3_policy05)[20,7])`, semi-partial _r_ = `r printnum(ch3_r_policy05_pun, digits=2)`), or abortion (_b_ = `r printnum(parameterEstimates(ch3_policy06)[20,4], digits=2)`, 95% CI [`r printnum(parameterEstimates(ch3_policy06)[20,8], digits=2)`, `r printnum(parameterEstimates(ch3_policy06)[20,9], digits=2)`], _p_ = `r printp(parameterEstimates(ch3_policy06)[20,7])`, semi-partial _r_ = `r printnum(ch3_r_policy06_pun, digits=2)`).

(ref:ch3policyPlotCoopCaption) _The cooperation latent variable predicts a host of economically progressive views._ For every Likert scale, 7 indicates increased support for the policy or agreement with the statement, whereas 1 indicates reduced support for the policy or disagreement with the statement. For visualisation ease, regression lines and 95% confidence interval shaded areas are predictions from least-squares linear regressions without socio-demographic covariates.

```{r ch3policyPlotCoop, echo=FALSE, warning=F, message=F, fig.width=7, fig.height=5.5, fig.cap="(ref:ch3policyPlotCoopCaption)"}
readd(ch3_policyPlotCoop)
```

### Discussion

The correlations, principal components analyses, and confirmatory factor analyses we report show that behaviour across a large suite of incentivised economic games can be captured by two underlying general preferences: cooperation and punishment. This finding replicates previous work [@Peysakhovich2014; @Reigstad2017] and demonstrates that the cooperative phenotype generalises to an English-speaking sample outside the United States and Europe. In addition, the unique combination of substantial monetary stakes, real-time interactive gameplay, and additional coordination games deployed in our study indicates that the cooperative phenotype emerges with methodologies that may more reliably track real-world social interactions and with games that do not involve exploitation of others [@Thielmann2020].

As predicted by the dual evolutionary framework of political ideology [@Claessens2020a], we found that the cooperative phenotype captured by our economic games was negatively related to SDO, with an effect size comparable to that found in a recent meta-analysis of personality traits and economic game behaviour [@Thielmann2020]. When we separately explored individual economic games, we found that the negative correlation between SDO and cooperation held for most cooperative decisions, corroborating previous work identifying negative correlations between SDO and cooperative behaviour in individual games [@Grunhage2020; @Haesevoets2015; @Halali2018; @Thielmann2020]. Further exploratory analyses revealed that the cooperative phenotype was related to economically progressive views across a wide range of policy issues, including environmentalism, welfare, taxation, and income redistribution. While the effect sizes for these associations were consistently small, the cooperation latent variable explained a comparable proportion of variance to other socio-demographic variables known to predict economically progressive policy views, such as age, gender, and education [@Feldman2014; @Malka2014], and the effects persisted when controlling for these variables.

```{r echo=F, warning=F, message=F}
loadd(ch3_compWorld01)
loadd(ch3_compWorld03)
```

We did not find the predicted association between the punishment latent variable and RWA, nor did we find any link between the punishment latent variable and socially conservative policy views. One explanation for this lack of relationship is that these punishment games do not reflect general group conformist preferences as expected. In fact, previous work [@Price2002; @Raihani2019] has argued that punishment in economic games may more closely track a competitive motive to increase one's relative payoffs over others, especially when the economic games are one-shot games with anonymous others in which there is no possibility for signalling, reputation building, or future behaviour modification. Given our pattern of results, there is reason to favour this competitive-punishment account over a normative-punishment account. We unexpectedly found a small positive relationship between SDO and the punishment latent variable, indicating that punishment in our games may be related to interpersonal competition rather than norm enforcement. To investigate this further, we conducted an additional analysis relating SDO and the cooperation and punishment latent variables to a view that we live in "a dog-eat-dog world where you have to be ruthless at times" [@Duckitt2009]. As expected, both SDO (_b_ = `r printnum(parameterEstimates(ch3_compWorld03)[19,4], digits=2)`, 95% CI [`r printnum(parameterEstimates(ch3_compWorld03)[19,8], digits=2)`, `r printnum(parameterEstimates(ch3_compWorld03)[19,9], digits=2)`], _p_ `r printp(parameterEstimates(ch3_compWorld03)[19,7])`) and the cooperation latent variable (_b_ = `r printnum(parameterEstimates(ch3_compWorld01)[19,4], digits=2)`, 95% CI [`r printnum(parameterEstimates(ch3_compWorld01)[19,8], digits=2)`, `r printnum(parameterEstimates(ch3_compWorld01)[19,9], digits=2)`], _p_ = `r printp(parameterEstimates(ch3_compWorld01)[19,7])`) were significantly related to a competitive worldview. Similarly, the punishment latent variable was significantly positively related to a competitive worldview (_b_ = `r printnum(parameterEstimates(ch3_compWorld01)[20,4], digits=2)`, 95% CI [`r printnum(parameterEstimates(ch3_compWorld01)[20,8], digits=2)`, `r printnum(parameterEstimates(ch3_compWorld01)[20,9], digits=2)`], _p_ = `r printp(parameterEstimates(ch3_compWorld01)[20,7])`), suggesting that punishment is reflecting a competitive motive rather than a norm-enforcing motive, at least in this instance.

In light of these concerns about the use of punishment games as a proxy for group conformist preferences, we conducted a second study in which we introduced two new tasks intended to measure two types of conformist preferences. The first task is designed to measure conformity in the form of social information use, while the second task is designed to measure conformity in the form of rule following. In Study 2, we re-sampled the same individuals from Study 1 for a second wave of data collection, and included these new tasks to further investigate the relationships between conformist preferences and RWA and social policy views.

## Study 2

### Methods

#### Participants

```{r echo=F, warning=F, message=F}
loadd(ch3_d2)
```

In Study 2, we attempted to contact every participant who had completed Study 1 and had not since withdrawn from our study or the larger New Zealand Attitudes and Values Study. This resulted in a sample frame with 997 participants. We did not call participants directly but instead emailed them with another invitation to participate in a similar study. Of the participants we emailed, 636 participants completed the second wave of data collection (64% retention rate). 27 of these participants did not complete the study in time or did not have RWA scores, resulting in a final sample of `r nrow(ch3_d2)` participants (`r sum(ch3_d2$Gender.T10 == 0, na.rm = TRUE)` females; age _M_ = `r printnum(mean(ch3_d2$Age.T10, na.rm = TRUE), digits = 0)` years, _SD_ = `r printnum(sd(ch3_d2$Age.T10, na.rm = TRUE), digits = 0)` years).

#### Materials

The materials for Study 2 were mostly identical to Study 1. As in Study 1, eight incentivised behavioural tasks were conducted online in real-time using oTree software [code available at https://osf.io/dwx8g/; @Chen2016]. The strategy method was used to elicit responses in all possible roles, and participants played for points which were converted to New Zealand dollars at the same rate as in Study 1 (1 point = $0.035). However, in Study 2, the Stag Hunt Game and the Stag Hunt Game with Punishment were replaced with two different tasks.

The first task was the Berlin Estimate AdjuStment Task [BEAST; @Molleman2019], a task designed to measure individual differences in conformity in the form of social information use. In this task, over five rounds, participants are required to report the number of animals present in images that are briefly displayed (see Appendix to Chapter 3, Figure \@ref(fig:ch3beastImages)). There are two parts to each round. In the first part, the image is displayed for six seconds and participants give their initial estimate of the number of animals ($E_1$). In the second part, participants are then made aware of a previous participant's estimate for the same image ($s$) and then give a second estimate ($E_2$). Previous participants' estimates are determined by searching a database of real estimates from 100 previous participants sampled by @Molleman2019. The computer searches for an estimate in this database that is (1) in the direction of the true (correct) answer (e.g., if the participant guesses 50 and the true answer is 60, social information will be higher than 50) and (2) deviates from the participant's first estimate according to a parameter $\Delta$: $s = E_1 * (1 \pm \Delta)$. Following the original paper, this $\Delta$ parameter varied across the five rounds, taking values of 0.25, 0.15, 0.20, 0.15, and 0.25 respectively. Participants were paid for their accuracy by randomly choosing a round and determining the distance between their estimate and the correct answer: 100 points - (distance * 5).

Based on participants' estimate in the first part, estimate in the second part, and the social information they were provided, we calculate a score for each of the five rounds using the following equation from @Molleman2019:

$$\text{BEAST score}=\frac{E_2 - E_1}{s - E_1}$$

In keeping with @Molleman2019, we remove cases where (1) participants moved their second estimate away from the social information, (2) participants moved their second estimate further than the social information, or (3) participants ran out of time to make their estimate (8% of cases). The BEAST score varies between 0 and 1, where 0 implies that the participant did not shift their estimate after receiving social information and 1 implies that the participant shifted their estimate completely towards the social information. This is our measure of social information use.

The second task was the Rule Following Task [@Kimbrough2018], a task designed to capture individual differences in conformity in the form of rule following. In this task, participants are required to place 30 balls, sequentially, into one of two buckets. Each ball placed into Bucket A earns the participant one point, while each ball placed into Bucket B earns the participant two points (randomly counterbalanced). Before making any decisions, participants are shown the text: "the rule is to place the balls into Bucket A". This costly exogenously-imposed rule is made clear on the screen before participants make any decisions. There is no penalty to participants who break the rule. Our measure of rule following is how often participants follow the rule by placing the balls into Bucket A.

#### Procedure

Data collection for Study 2 was conducted between 19^th^ October and 11^th^ November 2020 with weekly staggered recruitment. Study sessions contained between 30 and 130 participants. Sessions proceeded in a similar fashion to Study 1, except that the six economic games (Dictator Game, Trust Game, Public Goods Game, Ultimatum Game, Third-Party Punishment Game, and Second-Party Punishment Game) were presented in a random order first, and then the new BEAST and Rule Following Task were presented in a random order at the end of the session. As in Study 1, participants in a session were then randomly matched in real-time following the eight tasks and the computer determined bonus payments.

Participants were paid a fixed \$20 New Zealand dollar show-up fee, plus a bonus payment of between \$10 and \$35 (_M_ = \$`r printnum(mean(ch3_d2$egame.Paid.T11 - 20))`, _SD_ = \$`r printnum(sd(ch3_d2$egame.Paid.T11 - 20))`) depending on the decisions of themselves and others. On average, participants took `r printnum(mean(ch3_d2$egame.SecsAll.T11 / 60), digits = 0)` minutes to complete the eight tasks (_SD_ = `r printnum(sd(ch3_d2$egame.SecsAll.T11 / 60))` mins, range = `r printnum(min(ch3_d2$egame.SecsAll.T11 / 60), digits = 0)` - `r printnum(max(ch3_d2$egame.SecsAll.T11 / 60), digits = 0)` mins).

#### Pre-registration

We pre-registered two hypotheses on the Open Science Framework on 28^th^ September 2020 (https://osf.io/7vqah/). First, we hypothesised that RWA would positively predict social information use in the BEAST. Second, we hypothesised that RWA would positively predict rule following in the Rule Following Task.

#### Statistical analysis

We pre-registered separate Bayesian multilevel models to analyse the data from both new tasks. For the BEAST, we fitted a Bayesian multilevel beta regression model with BEAST score as the outcome variable. Beta regression is a useful technique for analysing continuous outcome variables that are bounded between 0 and 1 [@Douma2019]. We included RWA as the sole fixed-effect predictor, and random intercepts for image types and participants. For the Rule Following Task, we fitted a Bayesian multilevel logistic regression model with binary rule following (0 = break rule, 1 = follow rule) in each round as the outcome variable, round number and RWA as fixed-effect predictors, and random intercepts and slopes of round number for participants. As additional exploratory analyses, we also fitted Bayesian ordinal regressions predicting policy views from BEAST scores and rule following frequencies. We used weakly-informative priors, and all models converged normally ($\hat{R}$ = 1).

All analyses were conducted in R version 4.0.2 [@R-base]. We used the _brms_ package for Bayesian multilevel models [@R-brms]. Figures were created with the _ggplot2_ [@R-ggplot2] and _cowplot_ [@R-cowplot] packages, and the reproducibility of all analyses was ensured by using the _drake_ package [@R-drake].

#### Ethical approval

Ethical approval for Study 2 was provided by the University of Auckland Human Participants Ethics Committee, as an amendment to the approval for Study 1 (ref: 021666).

### Results

```{r echo=F, warning=F, message=F}
loadd(ch3_kfoldBEASTcompare)
s1 <- summary(readd(ch3_beast1))
s2 <- summary(readd(ch3_beast2))
```

Replicating @Molleman2019, we found that BEAST scores were skewed towards zero, with most participants choosing to stick with their initial estimate rather than shift their estimate towards the social information (Appendix to Chapter 3, Figure \@ref(fig:ch3histBEAST)). Nevertheless, there was also substantial variation in BEAST scores. In line with our first pre-registered hypothesis, we found that RWA positively predicted social information use in the BEAST. In a Bayesian multilevel beta regression analysis predicting BEAST scores (0 = no shift in estimate, 1 = complete shift in estimate to social information), the 95% credible interval for the RWA slope was greater than zero (_b_ = `r printnum(s1$fixed["RWA.T10.c","Estimate"],digits=2)`, 95% credible interval [`r printnum(s1$fixed["RWA.T10.c","l-95% CI"],digits=2)`, `r printnum(s1$fixed["RWA.T10.c","u-95% CI"],digits=2)`]; Figure \@ref(fig:ch3rfBeastPlot)a). This result held when additionally controlling for SDO, age, gender, ethnicity, education, socio-economic status, and local deprivation, individually and together in a full model (Appendix to Chapter 3, Table \@ref(tab:ch3beastTable)). In contrast, SDO did not predict social information use in the BEAST (_b_ = `r printnum(s2$fixed["SDO.T10.c","Estimate"],digits=2)`, 95% CI [`r printnum(s2$fixed["SDO.T10.c","l-95% CI"],digits=2)`, `r printnum(s2$fixed["SDO.T10.c","u-95% CI"],digits=2)`]).

(ref:ch3rfBeastPlotCaption) _Right-wing authoritarianism positively predicts social information use in the BEAST (a) but does not predict rule following in the Rule Following Task (b)._ Lines are mean posterior regression lines from Bayesian multilevel models without socio-demographic covariates, shaded areas are 95% credible intervals.

```{r ch3rfBeastPlot, echo=FALSE, warning=F, message=F, fig.width=7, fig.height=4, fig.cap="(ref:ch3rfBeastPlotCaption)"}
readd(ch3_rfBeastPlot)
```

```{r echo=F, warning=F, message=F}
s04 <- summary(readd(ch3_policyBEAST04))
s05 <- summary(readd(ch3_policyBEAST05))
s06 <- summary(readd(ch3_policyBEAST06))
s10 <- summary(readd(ch3_policyBEAST10))
s11 <- summary(readd(ch3_policyBEAST11))
s12 <- summary(readd(ch3_policyBEAST12))
s13 <- summary(readd(ch3_policyBEAST13))
```

BEAST scores also reliably predicted socially conservative views across a wide range of political issues (Figure \@ref(fig:ch3policyPlotBEAST); Appendix to Chapter 3, Tables \@ref(tab:ch3beastPolicyTable04) - \@ref(tab:ch3beastPolicyTable13)). Controlling for socio-demographics, exploratory Bayesian ordinal regressions revealed that mean BEAST scores (averaged over all five rounds) positively predicted support for religious instruction in schools (_b_ = `r printnum(s11$fixed["bScore","Estimate"],digits=2)`, 95% CI [`r printnum(s11$fixed["bScore","l-95% CI"],digits=2)`, `r printnum(s11$fixed["bScore","u-95% CI"],digits=2)`]), agreement with the essentialist belief that "a person's race biologically determines their abilities" (_b_ = `r printnum(s12$fixed["bScore","Estimate"],digits=2)`, 95% CI [`r printnum(s12$fixed["bScore","l-95% CI"],digits=2)`, `r printnum(s12$fixed["bScore","u-95% CI"],digits=2)`]), and disagreement with homosexuality being accepted by society (_b_ = `r printnum(s13$fixed["bScore","Estimate"],digits=2)`, 95% CI [`r printnum(s13$fixed["bScore","l-95% CI"],digits=2)`, `r printnum(s13$fixed["bScore","u-95% CI"],digits=2)`]). BEAST scores also negatively predicted support for abortion, both for any reason (_b_ = `r printnum(s06$fixed["bScore","Estimate"],digits=2)`, 95% CI [`r printnum(s06$fixed["bScore","l-95% CI"],digits=2)`, `r printnum(s06$fixed["bScore","u-95% CI"],digits=2)`]) and specifically when the woman's life is in danger (_b_ = `r printnum(s10$fixed["bScore","Estimate"],digits=2)`, 95% CI [`r printnum(s10$fixed["bScore","l-95% CI"],digits=2)`, `r printnum(s10$fixed["bScore","u-95% CI"],digits=2)`]). The only social issues not predicted by BEAST scores were support for same-sex marriage (_b_ = `r printnum(s04$fixed["bScore","Estimate"],digits=2)`, 95% CI [`r printnum(s04$fixed["bScore","l-95% CI"],digits=2)`, `r printnum(s04$fixed["bScore","u-95% CI"],digits=2)`]) and euthanasia (_b_ = `r printnum(s05$fixed["bScore","Estimate"],digits=2)`, 95% CI [`r printnum(s05$fixed["bScore","l-95% CI"],digits=2)`, `r printnum(s05$fixed["bScore","u-95% CI"],digits=2)`]), although these effects were in the expected direction. In contrast, BEAST scores did not predict any of the economic views predicted by the cooperative phenotype in Study 1.

(ref:ch3policyPlotBEASTCaption) _BEAST scores predict a host of socially conservative views._ For every Likert scale, 7 indicates increased support for the policy or agreement with the statement, whereas 1 indicates reduced support for the policy or disagreement with the statement. Regression lines and 95% credible interval shaded areas are predictions from Bayesian ordinal models including socio-demographic covariates.

```{r ch3policyPlotBEAST, echo=FALSE, warning=F, message=F, fig.width=7, fig.height=7, fig.cap="(ref:ch3policyPlotBEASTCaption)"}
readd(ch3_policyPlotBEAST)
```

```{r echo=F, warning=F, message=F}
loadd(ch3_corRF)
loadd(ch3_kfoldRFcompare)
s1 <- summary(readd(ch3_rf1))
s2 <- summary(readd(ch3_rf2))
s13 <- summary(readd(ch3_policyRF13))
```

Turning to the Rule Following Task, we found that rule following was bimodally distributed, with most participants either always following the rule or always breaking it (Appendix to Chapter 3, Figure \@ref(fig:ch3histRF)). Contrary to our second pre-registered hypothesis, we found that RWA did not predict rule following. The Spearman's rank correlation between RWA and the number of rounds in which participants followed the rule was `r printnum(ch3_corRF$estimate)` (_p_ = `r printp(ch3_corRF$p.value)`). Similarly, in a Bayesian multilevel logistic regression analysis predicting the probability of following the rule, the 95% credible interval for the log-odds RWA slope crossed zero (_b_ = `r printnum(s1$fixed["RWA.T10.c","Estimate"],digits=2)`, 95% CI [`r printnum(s1$fixed["RWA.T10.c","l-95% CI"],digits=2)`, `r printnum(s1$fixed["RWA.T10.c","u-95% CI"],digits=2)`]; Figure \@ref(fig:ch3rfBeastPlot)b). This result held when additionally controlling for all socio-demographics (Appendix to Chapter 3, Table \@ref(tab:ch3rfTable)). Interestingly, these additional models revealed that SDO negatively predicted the probability of rule following (_b_ = `r printnum(s2$fixed["SDO.T10.c","Estimate"],digits=2)`, 95% CI [`r printnum(s2$fixed["SDO.T10.c","l-95% CI"],digits=2)`, `r printnum(s2$fixed["SDO.T10.c","u-95% CI"],digits=2)`]), suggesting that high SDO participants were more likely to break the rule for personal gain. In further exploratory analyses, we found that the frequency of rule following predicted socially _progressive_ acceptance of homosexuality in society (_b_ = `r printnum(s13$fixed["rf","Estimate"],digits=2)`, 95% CI [`r printnum(s13$fixed["rf","l-95% CI"],digits=2)`, `r printnum(s13$fixed["rf","u-95% CI"],digits=2)`]) but did not predict any other social policy views.

### Discussion

In line with our first pre-registered hypothesis, we found that RWA, but not SDO, predicted increased conformity as measured by social information use in the BEAST. This builds on previous work with this task showing that social information use is associated with collectivist values [@Molleman2019]. Furthermore, we found that individual differences in social information use predicted socially conservative views across a wide range of political issues, such as abortion, homosexuality, and religious instruction in schools, but did not predict economic views. By relating RWA and social policy views to conformist preferences, these findings provide further support for the dual evolutionary framework of political ideology.

As with our findings with the norm-enforcing punishment games, we found no relationship between RWA and rule following, but we did find a relationship with SDO. One explanation for this is that, while intended to track willingness to follow a rule, the Rule Following Task also reflects elements of competitiveness, selfishness, and striving for personal gain: by breaking the rules, individuals can earn more money for themselves. The same can be said for the punishment games in Study 1: by removing points from others, individuals can competitively increase their relative earnings. The Rule Following Task also exogenously imposes a rule, but from the instructions alone it is unclear who has imposed the rule or for what reason. We might have found a different pattern of results if participants had been told that the rule had been imposed by another participant or by the consensus of their in-group.

## General discussion

Across two pre-registered studies, we found that behaviour in incentivised tasks was related to the two dimensions of political ideology and a wide variety of policy views. Participants who cooperated more in economic games were lower in SDO and expressed economically progressive policy views. Cooperation was unrelated to RWA or social views. Likewise, participants who conformed more to social information were higher in RWA and expressed socially conservative policy views. Social information use was unrelated to SDO or economic views. However, we also found that punishment and rule following behaviour did not show the predicted relationship with RWA or socially conservative policy views, and instead covaried with SDO. Taken together, these associations support and refine the general claim made by the dual evolutionary framework that the two dimensions of political ideology reflect phenotypic variation in preferences for cooperation and conformity.

We found that SDO and economic policy views were related to a cooperative phenotype latent variable estimated from multiple cooperation games. SDO was also related to most individual cooperative decisions, including decisions reflecting trust, altruism, coordination, and collective action. Recent work has characterised lower SDO and economically progressive views as specifically related to a reduced willingness to exploit others for personal gain [@Thielmann2020] and greater interpersonal trust [@Daniele2015]. Our exploratory analyses revealed that SDO is indeed related to exploitative behaviour, including returning less in the Trust Game, but it is also related to less cooperative behaviour in games that do not involve a conflict of interest and so do not allow for exploitation of others, such as not coordinating to mutual benefit in the Stag Hunt Game. Together, these findings support the prediction from the dual evolutionary framework that low SDO and support for economically progressive policies reflect a _general_ preference for cooperation that shapes people's views across a myriad of different social dilemmas that emerge from human group living, such as welfare benefit programmes, taxation systems, and collective responses to climate change.

We found that RWA and social policy views were related to conformity in the form of social information use. In contrast, RWA and social policy views were unrelated to norm-enforcing punishment or rule following. Taking these results at face value, this suggests that authoritarian and socially conservative views are perhaps better characterised by informational conformity than normative conformity [@Deutsch1955]. Informational conformity functions to adapt individual behaviour to local conditions by learning from peer behaviour, whereas normative conformity functions to minimise peer disapproval via enforcement of and adherence to social norms [@Claidiere2012]. A link between RWA and informational conformity fits with research showing that conservatives are more likely to share similar judgements and attitudes with like-minded people, exaggerate informational consensus within groups, and create echo chambers that promote the spread of misinformation [@Jost2018]. This link also fits with prior work linking social information use to collectivist values [@Molleman2019]. The finding that normative conformity does not differ across the political spectrum aligns with research showing that conservatives and liberals are equally likely to enforce their group's social norms and exhibit intolerance towards out-groups with different norms [@Brandt2014]. If this characterisation of authoritarian and socially conservative ideology is correct, we will need to refine the dual evolutionary framework of political ideology to place less emphasis on group-wide social norms and more emphasis on in-group informational conformity and social information use.

Nevertheless, it is also possible that our norm-enforcing punishment and rule following tasks were poor measures of a more general group conformist preference. These tasks contain competitive and selfish elements that may explain their correlations with SDO rather than RWA: punishment increases participants' relative earnings above others, and rule breaking earns more money. Moreover, these tasks did not explicitly measure social norms [@Krupka2013], did not contain any differentiation between in-groups and out-groups, and were one-shot rather than repeated, limiting the potential for the emergence and enforcement of group norms. Future research could adapt the methods used here to further refine our understanding of the nature of the conformist dimension of political ideology.

There are several important limitations of this work that should guide future research. First, while we collected data from a large, diverse, and representative adult sample (Appendix to Chapter 3, Figure \@ref(fig:ch3samplePlot)), it remains unclear to what extent our findings apply outside of New Zealand, both in other developed countries and in non-WEIRD [@Henrich2010] and small-scale societies. Future work should replicate these results in other populations around the world, including using adapted behavioural measures to predict locally-relevant political attitudes in small-scale societies. Second, the effect sizes in both studies were relatively small. However, smaller effect sizes are to be expected from relationships between behaviour and self-reported attitudes [@Dang2020] and do not preclude these relationships from being consequential in the long run [@Funder2019]. Third, one might argue that the reported associations between behaviour and political views could have simply arisen from people mapping their prior political beliefs onto the behavioural tasks. We believe this is unlikely, since the tasks were abstract, contained no political framing, and were not completed alongside ideology measures. Future work examining longitudinal data could expand on our cross-sectional study to formally test the directions of causality between general preferences and political attitudes.

In sum, we have shown clear differences in behavioural preferences across the two dimensions of political ideology. We found that a general cooperative preference covaried with SDO and economic policy views, and informational conformity covaried with RWA and social policy views. These psychological differences are in line with the predictions of the dual evolutionary framework of political ideology [@Claessens2020a]. Our results challenge claims that there are no fundamental psychological differences between liberals and conservatives [@Bakker2020; @Brandt2014; @Crawford2017a; @Crawford2017b], adding to the catalogue of traits that have been found to differ across the political landscape [@Jost2003; @Hibbing2014]. Moreover, by characterising the very different behavioural preferences underlying both ideological dimensions, our findings highlight the importance of work that moves beyond unidimensional scales to elucidate the dual evolutionary foundations of political ideology.

```{r echo=F, results='hide'}
# free up memory
rm(ch3_d1, ch3_d2, ch3_cor, ch3_pca1.2, ch3_pca2.2, ch3_cfa1, ch3_cfa2, ch3_sem1, 
   ch3_r_SDO_coop, ch3_r_SDO_pun, ch3_r_RWA_coop, ch3_r_RWA_pun, ch3_r_indGamesSDO, rs, 
   ch3_policy01, ch3_policy02, ch3_policy03, ch3_policy04, ch3_policy05, ch3_policy06, 
   ch3_policy07, ch3_policy08, ch3_policy09, ch3_r_policy01_coop, ch3_r_policy02_coop, 
   ch3_r_policy03_coop, ch3_r_policy07_coop, ch3_r_policy08_coop, ch3_r_policy09_coop, 
   ch3_r_policy04_pun, ch3_r_policy05_pun, ch3_r_policy06_pun, ch3_compWorld01, 
   ch3_compWorld03, ch3_kfoldBEASTcompare, s1, s2, s04, s05, s06, s10, s11, s12, 
   s13, ch3_corRF, ch3_kfoldRFcompare)
gc()
```

