# Behavioural preferences predict political party support beyond socio-demographic variables

## Introduction

```{r echo=F}
loadd(ch4_d)
```

Political parties form the bedrock of democratic societies, and disagreements between their adherents and detractors fuel the fire of modern politics. Political party support is a major predictor of policy views [@Cohen2003; @Leeper2014], partisan bias [@Bartels2002; @Slothuus2010], and voting [@Chen2019], and party affiliation is relatively stable over time [@Goren2005; @Green1994; @Schickler1997]. Given the importance of party support for political behaviour in the modern electorate, a central aim of the social sciences has been to understand why people gravitate towards some political parties over others [@Campbell1960; @Greene1999].

One theoretical model of political party support is the _domain-specific_ model [@Petersen2015; @Weeden2014; @Weeden2016; @Weeden2017]. Under this model, people support political parties that best advance their inclusive fitness interests [i.e., the adaptive goals of themselves and close family members; @Weeden2014; @Weeden2017]. Using socio-demographic data from the US General Social Survey, @Weeden2014 argued that wealthy religious people tend to support the Republican Party because Republicans promote policies that oppose income redistribution and sanction promiscuous behaviour that violates traditional religious norms. In contrast, minority ethnic groups tend to support the Democratic Party because Democrats promote race-based affirmative action policies. Similar effects of socio-economic index, religiosity, and ethnicity on political party support have emerged in other Western countries, such as New Zealand [@Greaves2017]. Based on socio-demographic findings like these, the domain-specific model claims that self-interest is a major determinant of political party support [@Weeden2017], echoing the seminal work of @Campbell1960 and @Converse1964.

An alternative theoretical model of political party support has been dubbed the _general-orientations_ model [@Weeden2016]. Under this model, people support political parties that best align with political ideologies that emerge from underlying psychological and behavioural differences [@Graham2009; @Haidt2012; @Hibbing2013; @Hibbing2014; @Jost2003; @Jost2009]. For example, research has found that, compared to Democratic Party supporters, Republican Party supporters tend to be more sensitive to disgusting stimuli [@Billingsley2018; @Shook2017] and threats to group status [@Mutz2018], and are less creative [@Tyagi2018], less open [@Rentfrow2009], and more conscientious [@Sweetser2014] in their personality profiles. Similar effects of personality on political party support have emerged in New Zealand [@Greaves2017]. The general-orientations model claims that fundamental psychological and behavioural differences manifest in a coherent political ideology which, alongside socio-demographic differences, predicts political party support [@Jost2003].

There has been much scholarly debate between proponents of the domain-specific and general-orientations models of political party support. On the one hand, @Hibbing2015 argues against the domain-specific model by noting that, while socio-demographic variables are indeed important predictors of political party support, people often support political parties whose policies run counter to their self-interest (e.g., wealthy people supporting the Democrats). On the other hand, researchers have argued against the general-orientations model by claiming that there exist no fundamental psychological or behavioural differences across the political landscape [@Greenberg2003], with traits like threat sensitivity, intolerance, and prejudice being prevalent among liberals and conservatives alike [@Bakker2020; @Brandt2014; @Crawford2017b]. In addition, @Weeden2016 show that economic and social policy views do not reliably covary in the US, a finding which they claim "undermine[s] views... that posit that deep and ancient psychological foundations push individuals into broadly coherent liberal or conservative political issue positions" (p. 54). Yet, despite this ongoing debate, no work to date has directly tested between these models by comparing the relative predictive performance of behavioural and socio-demographic variables for explaining political party support.

Here, we conduct such a test. In addition to a host of socio-demographic variables, we focus on two behavioural variables highlighted by recent theoretical work: behavioural preferences for cooperation and group conformity. The dual evolutionary framework of political ideology [@Claessens2020a] argues that, rather than varying along a single liberal-conservative spectrum, political ideology is organised along two dimensions shaped by two key trade-offs that naturally emerge from human group living: a trade-off between cooperative egalitarianism and competitive hierarchy, and another trade-off between individualism and group conformity. The resulting dimensions of political ideology have been variously labelled economic and social conservatism [@Feldman2014], social dominance and authoritarianism [@Duckitt2009], and individualising and binding foundations [@Haidt2012]. This two-dimensional general-orientations framework is perfectly compatible with previous findings showing that economic and social policy views are uncorrelated [@Weeden2016], since both ideological dimensions are motivated by distinct behavioural preferences.

One key prediction of the dual evolutionary framework is that variation in behavioural preferences for cooperation and group conformity should predict political ideology, policy views, and political party support in modern humans. In line with this, @Claessens2020b found that behavioural preferences for cooperation and conformist "social information use" covaried with two widely-used measures of political ideology, social dominance orientation and right-wing authoritarianism, respectively. Moreover, cooperative and conformist preferences predicted a variety of economic and social policy views, respectively. However, this study did not examine the relationships between cooperative and conformist preferences and political party support, leaving the link between behavioural preferences and real-world voting intentions unclear.

Previous research examining the link between cooperative and conformist preferences and political party support is mixed and sparse. While some studies have found that cooperative preferences predict support for centre-left political parties in Europe [@Cappelen2017; @Grunhage2020; @Kerschbamer2020; @Muller2019], studies in the United States have found no relationship between cooperative preferences and support for Republicans or Democrats [@Anderson2005; @Balliet2016]. Regardless, this work did not directly test between the domain-specific and general-orientations models of political party support by comparing the relative importance of behavioural preferences and socio-demographic variables. Moreover, no research to date has examined the relationship between conformist "social information use" and political party support, despite the promising link between social information use and socially conservative and authoritarian political attitudes [@Claessens2020b].

In order to test between the domain-specific and general-orientations models of political party support, here we compare the relative importance of cooperative and conformist behavioural preferences and socio-demographic variables for predicting political party support. We leverage nationally-representative survey data capturing socio-demographics and reported support for the main political parties in New Zealand, and behavioural data capturing cooperative and conformist preferences in a series of incentivised tasks. In particular, we focus on several socio-demographic variables highlighted by previous work as important predictors of political party support: age, gender, ethnicity, education, socio-economic status, and religiosity [@Greaves2017; @Weeden2014; @Weeden2016; @Weeden2017]. The domain-specific model hypothesises that these socio-demographics will be more important predictors of political party support than behavioural preferences, since these variables best capture inclusive fitness interests [@Weeden2014]. In contrast, the general-orientations model and the dual evolutionary framework hypothesise that behavioural preferences will be important predictors of political party support, beyond socio-demographic variables [@Claessens2020a].

## Methods

### Participants and sampling

Participants were sampled from the New Zealand Attitudes and Values Study (NZAVS), an annual longitudinal self-report study of the New Zealand population that has been active since 2009. Participants are sampled for the NZAVS through random draws from the New Zealand electoral roll. For the incentivised behavioural tasks, we initially contacted 3345 NZAVS participants who had previously expressed an interest in participating in additional studies (see Appendix to Chapter 3, Supplementary Methods for sample frame criteria). Of these 3345 participants, 1043 participated in our first wave of incentivised behavioural tasks in 2019, were successfully paid for the tasks, and did not time out of the study. 631 of these participants went on to successfully complete our second wave of incentivised behavioural tasks in 2020. For the purposes of this study, we focus on behavioural data from the second wave of incentivised behavioural tasks, since this wave contained both cooperation tasks and the social information use task (_N_ = `r nrow(ch4_d)`; `r sum(ch4_d$Gender.T11 == "Female", na.rm = T)` females; mean age = `r printnum(mean(ch4_d$Age.T11, na.rm = T), digits = 0)` years, age range = `r printnum(min(ch4_d$Age.T11, na.rm = T), digits = 0)` - `r printnum(max(ch4_d$Age.T11, na.rm = T), digits = 0)` years). In line with the results of the New Zealand 2020 General Election, our sample was slightly left-leaning politically, showing more support for progressive political parties than conservative political parties (Appendix to Chapter 4, Figure \@ref(fig:ch4plotHist)).

### Materials

#### New Zealand Attitudes and Values Study measures

We used self-report variables from the 11^th^ wave of the NZAVS (data collected from 2018 to 2019). Our main outcome variable was political party support measured with a 7-point Likert scale. Participants were asked "please rate how strongly you oppose or support each of the following political parties", with 1 indicating "strongly oppose" and 7 indicating "strongly support". Participants answered this question for the primary political parties in New Zealand: the National Party, the Labour Party, the Green Party, the ACT Party, the New Zealand First Party, and the M&#257;ori Party. To get a sense of where these parties sat on various economic and social policies around the time of data collection, we accessed the popular website @PoliticalCompass2020 which, in advance of the New Zealand 2020 General Election, had estimated the placement of the major New Zealand political parties on the two dimensions of political ideology. This website estimated that the ACT Party had the most economically conservative policies, followed by (in descending order) National, New Zealand First, Labour, M&#257;ori, and Greens. The same website estimated that New Zealand First had the most socially conservative policies, followed by (in descending order) National, Labour, M&#257;ori, ACT, and Greens.

We also included several socio-demographic variables from the NZAVS that have previously been shown to predict political party support [@Greaves2017; @Weeden2014; @Weeden2016; @Weeden2017]. These included gender, age, education, ethnicity, religiosity, socio-economic status indexed by occupational role, and local deprivation level (see Appendix to Chapter 4, Table \@ref(tab:ch4itemTable)).

#### Incentivised behavioural tasks

As described in Chapter 3, participants completed eight incentivised behavioural tasks online in real-time using oTree [@Chen2016]. Three of these tasks were economic games measuring cooperation, three tasks were economic games measuring norm-enforcing punishment, one task measured rule following, and one task measured social information use [@Claessens2020b]. For the purposes of this study, we focus on the cooperation and social information use tasks, since these measures have been shown to relate to political ideology and policy views in a previous study [@Claessens2020b]. In all tasks, participants were paid in points that were converted to New Zealand dollars at a rate of 1 point = $0.035.

We used the following cooperation games that require participants to incur a personal cost to benefit others:

- _Dictator Game_. Player A is given 100 points. They must decide how many of these points to transfer to Player B. Player A keeps the remaining points. Player B is passive in the interaction.
- _Trust Game_. Players A and B both start with 50 points. First, Player A decides whether or not to transfer all 50 points to Player B, in the knowledge that the transferred amount will be tripled to 150 points. If Player A transfers, Player B now has 200 points. Player B must then decide to transfer between 0 and 150 points back to Player A.
- _Public Goods Game_. Four players begin with 100 points each. They can contribute between 0 and 100 points into a shared group project. All four decisions are made simultaneously, and then the amount in the group project is doubled and distributed evenly between all four players. Each player ends the game with their share from the group project, plus the points they initially refrained from contributing.

We also employed the Berlin Estimate AdjuStment Task to measure social information use [BEAST; @Molleman2019]. In this task, participants had to estimate the number of animals in several briefly-displayed images (see Appendix to Chapter 3, Figure \@ref(fig:ch3beastImages)). In each of five rounds, participants saw an image for six seconds and provided an estimate of the number of animals. Participants were then shown a previous participant's estimate of the number of animals (social information) and could modify their initial estimate based on this new information. Social information was drawn from a real dataset of previous estimates from @Molleman2019. The software chose a previous estimate that (1) was in the direction of the true answer, and (2) deviated from the participant's initial estimate according to a parameter that varied across rounds [@Claessens2020b]. Participants were paid by drawing a randomly-chosen estimate and determining the accuracy of that estimate: 100 points - (distance * 5).

From the participant's initial estimate ($E_1$), the social information they received ($s$), and their second estimate ($E_2$), we calculated a "BEAST score" for each round following the equation in @Molleman2019:

$$\text{BEAST score}=\frac{E_2 - E_1}{s - E_1}$$

We removed cases in which (1) participants moved their second estimate in the opposite direction from the social information, (2) participants moved their second estimate further than the social information, or (3) participants ran out of time to make their estimate (8% of cases). The BEAST score varies continuously between 0 and 1, where 0 indicates less reliance on social information and 1 indicates complete conformity to social information.

### Procedure

Participants completed the 11^th^ wave of the NZAVS self-report survey between 2018 and 2019. Participants completed the incentivised behavioural tasks in October and November 2020. Online real-time study sessions contained between 30 and 130 participants. In each session, participants completed the cooperation and punishment games in a randomised order first, followed by the social information use and rule following tasks in a randomised order. After completing the tasks, participants were then randomly matched in real-time and the computer determined bonus payments. Participants were paid a fixed \$20 New Zealand dollar show-up fee, plus a bonus payment of between \$10 and \$35 (_M_ = \$`r printnum(mean(ch4_d$egame.Paid.T11-20))`, _SD_ = \$`r printnum(sd(ch4_d$egame.Paid.T11-20))`) depending on the decisions of themselves and others. On average, participants took `r printnum(mean(ch4_d$egame.SecsAll.T11 / 60),digits=0)` minutes to complete the eight tasks (_SD_ = `r printnum(sd(ch4_d$egame.SecsAll.T11 / 60))` mins, range = `r printnum(min(ch4_d$egame.SecsAll.T11 / 60),digits=0)` - `r printnum(max(ch4_d$egame.SecsAll.T11 / 60),digits=0)` mins).

### Statistical analysis

```{r echo=F}
loadd(ch4_cfa)
loadd(ch4_m0)
```

To deal with missing data, we employed multiple imputation with predictive mean matching for continuous variables and logistic regression for binary variables [@vanBuuren2018]. We imputed ten datasets, based on the recommended best practice of imputing a number of datasets equal to the average percentage of missing data across variables (8% in our dataset). Visual inspection of the imputed variables suggested that the imputations were reasonable (Appendix to Chapter 4, Figure \@ref(fig:ch4plotImpute)).

In preparation for our main analysis, we constructed two variables to capture participants' cooperative and conformist behavioural preferences, using a similar approach as in previous work [@Claessens2020b]. For the cooperation variable, we fitted a confirmatory factor model with loadings from all the cooperative decisions in the economic games (RMSEA = `r printnum(fitMeasures(ch4_cfa)['rmsea.scaled'])`; SRMR = `r printnum(fitMeasures(ch4_cfa)['srmr'])`; CFI = `r printnum(fitMeasures(ch4_cfa)['cfi.scaled'])`) and extracted latent factor scores. For the conformity variable, we calculated average BEAST scores across the five rounds.

For our main analysis, we fitted Bayesian multilevel cumulative link regression models to the multiply-imputed datasets. We used cumulative link models to deal with the ordinal nature of the Likert scale "political party support" outcome variable. We analysed the dataset in long-form, whereby each participant had six rows in the dataset, one for each political party (total number of rows = `r printnum(nrow(ch4_m0$data), digits = 0)`). We included random intercepts for participants, and random intercepts and slopes for political parties. First, we fitted two models that aligned with the verbal claims from the domain-specific and general-orientations accounts: (1) a model where all socio-demographics predict party support but behavioural preferences do not (i.e., domain-specific model), and (2) a model where all socio-demographics _and_ behavioural preferences predict party support (i.e., general-orientations model). Next, we fitted models with each socio-demographic and behavioural preference as a separate predictor, to determine their individual effects on political party support. Finally, we fitted a null intercept-only model as a lower-end baseline for comparison. Prior predictive simulation was used to choose weakly-informative priors that gave equal prior probability to each ordinal outcome [@McElreath2020]. In all models, the ordinal education variable was modelled as a monotonic predictor [@Burkner2020]. All models converged normally ($\hat{R}$ = 1). The predictive performance of each model on out-of-sample data was compared with approximate leave-one-out cross-validation using Pareto smoothed importance sampling [@R-loo]. In particular, we compared differences in the expected log pointwise predictive density (ELPD) between models, with a higher ELPD indicating better out-of-sample predictive performance.

All analyses were conducted in R version 4.0.2 [@R-base]. We used the _mice_ package for multiple imputation [@R-mice], the _lavaan_ package to construct the confirmatory factor model [@R-lavaan], the _brms_ package for Bayesian multilevel cumulative link regression models [@R-brms], and the _loo_ package for approximate leave-one-out cross validation [@R-loo]. Figures were created with the _ggplot2_ [@R-ggplot2] and _cowplot_ [@R-cowplot] packages, and the reproducibility of all analyses was ensured by using the _drake_ package [@R-drake].

### Ethical approval

Ethical approval was provided by the University of Auckland Human Participants Ethics Committee on 19^th^ July 2018 for three years (ref: 021666).

## Results

```{r echo=F}
loadd(ch4_looDiffFull)
```

Model comparison revealed that the general-orientations model outperformed the domain-specific model (ELPD difference = `r printnum(abs(ch4_looDiffFull[2,1]))`, standard error = `r printnum(ch4_looDiffFull[2,2])`; Figure \@ref(fig:ch4plotLOO)a-\@ref(fig:ch4plotLOO)c). In other words, the out-of-sample predictive accuracy of the model including only socio-demographic predictors (i.e., domain-specific model) improved when additionally including information about participants' behavioural preferences (i.e., general-orientations model). Further comparison of models with individual predictors revealed that this improvement was driven by participants' cooperative preferences, rather than their conformist preferences. All socio-demographic and behavioural variables, except BEAST scores and ethnicity, improved out-of-sample prediction of political party support compared to the null intercept-only model (Figure \@ref(fig:ch4plotLOO)d). The only socio-demographic variable that improved out-of-sample prediction beyond the cooperation model was education (Figure \@ref(fig:ch4plotLOO)e). Models including other socio-demographics performed no better than the cooperation model, indicating that these variables are no more predictive of out-of-sample political party support than the "cooperative phenotype".

(ref:ch4plotLOOCaption) _Model comparison supports the general-orientations model over the domain-specific model._ (_a_) Visual depiction of the general-orientations model. (_b_) Visual depiction of the domain-specific model. (_c_) Improvement in expected log pointwise predictive density (ELPD) for the general-orientations model compared to the domain-specific model. (_d_) Improvement in ELPD for models with individual predictors compared to the null intercept-only model. (_e_) Improvement in ELPD for models with individual predictors compared to the cooperation model. Lines represent +/- 2 standard errors.

```{r ch4plotLOO, echo=FALSE, warning=F, message=F, fig.width=7, fig.height=7, fig.cap="(ref:ch4plotLOOCaption)"}
readd(ch4_plotLOO)
```

```{r echo=F}
loadd(ch4_slopeLabour1)
loadd(ch4_slopeNational1)
loadd(ch4_slopeGreens1)
loadd(ch4_slopeACT1)
loadd(ch4_slopeNZFirst1)
loadd(ch4_slopeMaori1)
loadd(ch4_beastNZFirst)
```

Posterior predictions from our multilevel models further revealed effects of behavioural preferences on support for the different political parties. We found that the cooperative phenotype was positively related to support for the Labour Party (median posterior slope = `r printnum(median(ch4_slopeLabour1))`, 95% credible interval [`r printnum(quantile(ch4_slopeLabour1, 0.025))` `r printnum(quantile(ch4_slopeLabour1, 0.975))`]) and the Green Party (median posterior slope = `r printnum(median(ch4_slopeGreens1))`, 95% CI [`r printnum(quantile(ch4_slopeGreens1, 0.025))` `r printnum(quantile(ch4_slopeGreens1, 0.975))`]) and was negatively related to support for the National Party (median posterior slope = `r printnum(median(ch4_slopeNational1))`, 95% CI [`r printnum(quantile(ch4_slopeNational1, 0.025))` `r printnum(quantile(ch4_slopeNational1, 0.975))`]) and the ACT Party (median posterior slope = `r printnum(median(ch4_slopeACT1))`, 95% CI [`r printnum(quantile(ch4_slopeACT1, 0.025))` `r printnum(quantile(ch4_slopeACT1, 0.975))`]; Figure \@ref(fig:ch4plotCoopSlopes); Appendix to Chapter 4, Figure \@ref(fig:ch4plotm1)). The cooperative phenotype was only weakly positively related to support for the M&#257;ori Party (median posterior slope = `r printnum(median(ch4_slopeMaori1))`, 95% CI [`r printnum(quantile(ch4_slopeMaori1, 0.025))` `r printnum(quantile(ch4_slopeMaori1, 0.975))`]), and was unrelated to support for the New Zealand First Party (median posterior slope = `r printnum(median(ch4_slopeNZFirst1))`, 95% CI [`r printnum(quantile(ch4_slopeNZFirst1, 0.025))` `r printnum(quantile(ch4_slopeNZFirst1, 0.975))`]). This pattern of results was robust to controlling for all socio-demographics in the full model. In contrast, we found that BEAST scores did not predict political party support in our primary multilevel models (Appendix to Chapter 4, Figures \@ref(fig:ch4plotBeastSlopes) and \@ref(fig:ch4plotm2)), and this result also held when controlling for all socio-demographics. It should be noted, however, that when we fitted individual fixed effects regressions analysing support for each political party separately, we found a weak positive effect of BEAST scores on support for the New Zealand First Party (median posterior slope = `r printnum(summary(ch4_beastNZFirst, robust = TRUE)$fixed["beast","Estimate"])`, 95% CI [`r printnum(summary(ch4_beastNZFirst, prob = 0.95)$fixed["beast","l-95% CI"])` `r printnum(summary(ch4_beastNZFirst, prob = 0.95)$fixed["beast","u-95% CI"])`]), but not for other political parties (Appendix to Chapter 4, Figure \@ref(fig:ch4plotBeast)).

(ref:ch4plotCoopSlopesCaption) _Cooperative phenotype predicts political party support._ Posterior slopes for the cooperative phenotype predicting support for New Zealand political parties, from the multilevel model without socio-demographic controls. Political parties are ranked from the most economically progressive (top) to the most economically conservative (bottom), as estimated by @PoliticalCompass2020. Densities represent posterior distributions, points represent medians, and lines represent 50% and 95% credible intervals.

```{r ch4plotCoopSlopes, echo=FALSE, warning=F, message=F, fig.width=5, fig.height=4, fig.cap="(ref:ch4plotCoopSlopesCaption)", dev='cairo_pdf'}
readd(ch4_plotCoopSlopes)
```

Individual socio-demographic models revealed effects on political party support broadly consistent with previous work [Appendix to Chapter 4, Figures \@ref(fig:ch4plotm3) - \@ref(fig:ch4plotm9); @Greaves2017]. Men reported greater support for National and ACT, but less support for Labour, Greens, and the M&#257;ori Party. Older participants reported greater support for National and ACT, but less support for Labour and Greens. Religious participants reported greater support for National, ACT, and New Zealand First, but less support for Labour and Greens. Participants in areas with higher local deprivation reported greater support for Labour, Greens, and New Zealand First, but less support for National and ACT. Educated participants reported greater support for Labour, Greens, and the M&#257;ori Party, but less support for National and ACT. Participants with high socio-economic index also reported greater support for Labour, Greens, and the M&#257;ori Party, but less support for National and ACT, a result which runs contrary to self-interest claims. Ethnicity was unrelated to political party support.

## Discussion

In a large sample of New Zealand voters, we found that behavioural preferences predicted political party support beyond socio-demographic variables. The "cooperative phenotype", estimated from a broad suite of incentivised economic games, positively predicted support for the Labour Party and the Green Party, and negatively predicted support for the National Party and the ACT Party. Conformist "social information use" did not predict political party support in our primary multilevel models, but did weakly predict support for the New Zealand First Party in a separate fixed-effects regression. Together, these findings suggest that the domain-specific model of political party support, which emphasises socio-demographic variables, is incomplete. Instead, in line with the general-orientations model and the dual evolutionary framework of political ideology, we have shown that fundamental differences in behavioural preferences are just as important for explaining political party support.

Our results support and build on previous work with this sample showing that the cooperative phenotype is negatively associated with social dominance orientation and positively associated with economically progressive policy views [@Claessens2020b]. Here, we find that the cooperative phenotype is related not only to self-reported attitudes but also to real-world voting intentions as measured by support for New Zealand political parties. In line with research in European countries [@Cappelen2017; @Grunhage2020; @Kerschbamer2020; @Muller2019], we found that cooperative preferences were positively associated with support for economically progressive political parties that promote income redistribution and taxes on the wealthy (e.g., the Green Party), and negatively associated with support for economically conservative political parties that promote free market capitalism (e.g., the ACT Party). By contrast, cooperative preferences did not predict support for the New Zealand First Party, who are relatively more centrist with their economic policies [@PoliticalCompass2020]. Thus, there is reason to believe that, in line with the dual evolutionary framework [@Claessens2020a], both our findings and results from previous work in Europe have emerged from an underlying association between cooperative preferences and party positions on _economic_ issues.

We did not find an association between social information use and support for political parties in our primary multilevel models. This is an unexpected finding, since a previous study with this sample revealed that behaviour in the BEAST task predicted right-wing authoritarianism and socially conservative policy views [@Claessens2020b]. One explanation for these differing results could be that, unlike in the United States, socially conservative voting intentions do not vary substantially across New Zealand, with most of the variation in political party support being explained by differences in economic policies. The relatively small effect of religion in our multilevel models perhaps supports this view.

Another explanation for the lack of association between social information use and political party support is that, in line with the rigidity-of-the-extreme hypothesis [@Zmigrod2020; @Greenberg2003], conformist behavioural preferences might be found at ideological extremes in both conservative _and_ liberal directions. Recent work on the existence of left-wing authoritarianism [@Costello2020; @Conway2018] highlights this as a very real possibility. Conformist preferences at the extremes of the ideological landscape would be consistent with research in the United States showing that partisanship towards both the Republican Party _and_ the Democrat Party is associated with cognitive inflexibility [@Zmigrod2020]. A similar "conformity-of-the-extreme" effect could potentially mask any relationships between social information use and political party support in our study. For example, some conformist individuals might show strong support for the National Party and less support for the Labour Party, while other conformist individuals might show the reverse pattern. Since these influences cancel one another out, using social information use as a single linear predictor of political party support reveals no effect overall. To overcome this methodological issue and test whether conformity is related to the extremity of political partisanship, future research could adopt the methods in @Zmigrod2020 and relate social information use to continuous "identity fusion" between different political parties.

Nevertheless, in separate fixed effects regressions, we did find that social information use weakly predicted support for New Zealand First, the most socially conservative political party. While this latter result is in line with the general-orientations model and the dual evolutionary framework of political ideology, the small effect should be interpreted with caution. Fixed effects models are less conservative in their estimation of parameter uncertainty and do not allow for generalisation beyond a particular sample and set of stimuli [@Yarkoni2020]. Multilevel models, on the other hand, offer both improved out-of-sample prediction [@McElreath2020] and the possibility of generalisation beyond an idiosyncratic sample and set of stimuli [@Yarkoni2020]. In order to generalise this statistical inference to further political parties and participants, future work could collect data from a larger, more politically diverse sample, perhaps also measuring support for minority New Zealand political parties with more politically-extreme stances, such as the New Conservative Party, the ONE Party, the Opportunities Party, and Vision NZ.

In line with previous research in New Zealand [@Greaves2017], we found that support for economically and socially conservative political parties was associated with being older, religious, male, and less locally deprived. Interestingly, people who were less educated and lower in socio-economic index (operationalised by occupational role) also supported more conservative parties, a finding that runs counter to the self-interest argument: these are the very people who would benefit most from progressive policies. In addition, we found no effect of ethnicity, possibly because our sample was skewed towards New Zealand Europeans (`r printnum((sum(ch4_d$EthnicCats.T11 == "Pakeha", na.rm = T) / nrow(ch4_d)) * 100, digits = 0)`% of the sample). That this pattern of results held in the full model including behavioural preferences suggests that both socio-demographic and behavioural variables have independent effects on political party support. This raises interesting questions about how socio-demographics and behavioural variables interact to predict political party support, especially when the two are in conflict. Would a cooperative person continue to support economically progressive political parties if they lived in a less deprived area? Would a competitive person continue to support economically conservative political parties if they were better educated? While beyond the scope of the current project, future research could systematically interact socio-demographic and behavioural variables in multilevel regression models to answer these questions.

In sum, we provide evidence in support of the general-orientations model of political party support, showing that behavioural preferences predict party support beyond socio-demographic factors. Importantly, our study is a conservative test of this model, drawing from just one particular general-orientations framework: the dual evolutionary foundations of political ideology [@Claessens2020a]. Future work should expand this approach to include the myriad of additional psychological and behavioural preferences that have been shown to vary across the political landscape, including personality [@Greaves2017], threat sensitivity [@Hibbing2014], and disgust sensitivity [@Shook2017]. We expect that this future research will further support our contention that, while self-interest is undoubtedly an important determinant of political attitudes, support for political parties and party positions themselves align with fundamental psychological and behavioural preferences.

```{r echo=F, results='hide'}
# free up memory
rm(ch4_d, ch4_cfa, ch4_m0, ch4_looDiffFull, ch4_slopeLabour1, ch4_slopeNational1, 
   ch4_slopeGreens1, ch4_slopeACT1, ch4_slopeNZFirst1, ch4_slopeMaori1, ch4_beastNZFirst)
gc()
```

