# Cooperation, social norms, and the dual dimensions of political ideology

```{r warning=F, message=F, echo=F}
loadd(ch6_d0, ch6_d, ch6_looDiff1, ch6_looDiff2, ch6_looDiff3, 
      ch6_post5, ch6_plotSDO, ch6_plotRWA, ch6_summary1, ch6_summary2a, ch6_summary3,
      ch6_summary4, ch6_summary5, ch6_summary7, ch6_summary8, ch6_summary9,
      ch6_summary11)
```

## Introduction

Humans disagree on a variety of political issues. Some people believe that income should be redistributed to those who need it most, while others believe that such redistribution is unfair to those who earned their wealth. Some people believe that traditional norms, such as heterosexual marriage and gender roles, should be preserved, while others believe that these antiquated norms should be abolished. These core beliefs motivate people to support different political parties and vote for different political candidates. A central research aim of the social sciences is to understand the psychology underlying people's political attitudes and values [@Campbell1960; @Converse1964].

Over the past decades, evidence from the social sciences has revealed that political attitudes are organised along two dimensions of political ideology [@Claessens2020a; @Feldman2014; @Duckitt2009; @Trompenaars1993; @Everett2013; @Ashton2005]. The first dimension, often referred to as economic conservatism or social dominance, captures views on free market capitalism, big business, economic inequality, and taxation. The second dimension, often referred to as social conservatism or authoritarianism, captures views on abortion, same-sex marriage, gender roles, and religious instruction in schools. Together, researchers argue, these two dimensions of political ideology better explain variation in political attitudes and values than a single liberal-conservative spectrum [@Duckitt2009].

The dual evolutionary foundations framework of political ideology explains the two-dimensional structure of ideology as arising naturally from the tensions that characterise human group living [@Claessens2020a]. In particular, the framework argues that the two dimensions of political ideology reflect two key trade-offs inherent to human groups: a trade-off between competitive hierarchy and cooperative egalitarianism, and a trade-off between individualism and group conformity to social norms. According to this theory, these trade-offs produce individual differences in both striving for personal gain and normative conformity, behavioural preferences which are reflected in two dimensions of political attitudes in modern humans [@Claessens2020a].

A recent study set out to test these predictions from the dual evolutionary foundations framework [@Claessens2020b]. In the study, a large sample of New Zealanders completed a battery of incentivised behavioural tasks measuring cooperative and conformist preferences. In support of the framework, the study found that cooperative preferences (e.g., altruism, trust, collective action) were related to social dominance orientation and a host of economic policy views, and conformist behaviour in the form of social information use was related to right-wing authoritarianism and a host of social policy views. However, other findings appeared to contradict the theory. Right wing authoritarianism and social policy views were unrelated to norm-enforcing punishment and rule following behaviours, raising questions about the relationship between authoritarianism and normative conformity.

There are several limitations with the tasks used to measure normative conformity in this previous work. First, the tasks did not explicitly measure social norms. Without such a measure, it is impossible to know whether rule following and punitive behaviours are actually attempts to adhere to and enforce perceived social norms. Second, the tasks do not adequately separate normative conformity from competitive preferences. Since @Claessens2020b found that punishment and rule breaking were related to social dominance orientation rather than right-wing authoritarianism, the authors suggested that these behaviours might reflect competitive rather than conformist motives. Punishment can be a competitive strategy, increasing one's own payoff relative to others [@Raihani2019], and rule breaking can be a selfish strategy, maximising one's payoff. Thus, in order to further test the relationship between normative conformity and the social dimension of ideology, we require an incentivised behavioural task that explicitly measures social norms and distinguishes between selfish and conformist preferences.

One such incentivised behavioural task has emerged from the literature on norm-dependent utility models in behavioural economics [@Kimbrough2018; @Krupka2009; @Krupka2013; @Kessler2012]. Norm-dependent utility models explain deviations from self-interest in humans. These models state that individuals are driven by their economic self-interest, but also receive negative utility from violating social norms. Under this view, if the social norms in a given situation prescribe prosocial behaviour and individuals are sensitive enough to social norms, then individuals may deviate from their economic self-interest [@Kimbrough2018]. For example, Lucy may choose not to litter in the park because there exist strong anti-littering norms and she has a particular proclivity for adhering to such norms. Interestingly, although norm-dependent utility models were devised independently from political psychology, these models have converged upon two dimensions of human sociality that are similar to those thought to underlie political ideology in humans: sensitivity to personal gain and sensitivity to normative standards of behaviour [@Claessens2020a].

Behavioural economists have developed empirical paradigms to test predictions of norm-dependent utility models. One such behavioural paradigm is "mini-Dictator Games" (henceforth mini-DGs). @Panizza2019 presented participants with a series of binary mini-DGs in which they had to decide between two different splits of points between themselves and another player. For example, in one mini-DG, the first option might split the points evenly between the two players, while the second option might allow the dictator to keep all the points to themselves. After completing a series of these binary choices, @Panizza2019 explicitly measured social norms by asking participants to rate how socially appropriate _the majority of other participants_ believed it was to choose each option [@Krupka2013]. Using both mini-DG decisions and social appropriateness ratings, @Panizza2019 then estimated two parameters for each participant which reflected the two components of norm-dependent utility: sensitivity to personal gain (the "payoff" utility component) and sensitivity to social norms (the "normative" utility component). The researchers found that these parameters both predicted choices in the mini-DGs, supporting norm-dependent utility models.

By explicitly measuring social norms and estimating separate parameters for both selfish and conformist preferences, the mini-DG paradigm represents a promising new tool for examining the dual foundations of political ideology. Here, we use the mini-DG paradigm with an online sample of US participants, and relate behaviour in the tasks to the two dimensions of political ideology. In particular, we measure social dominance orientation (SDO) and right-wing authoritarianism [RWA; @Duckitt2009], widely-used two-dimensional measures of political ideology. SDO is a measure of group-based dominance and anti-egalitarianism, while RWA is a measure of submission to authority, conformity, and authoritarian aggression towards outgroups and deviants. We predict that SDO will be related to individual differences in sensitivity to personal gain (the payoff utility component), whereas RWA will be related to individual differences in adherence to social norms (the normative utility component).

## Methods

### Power analysis

We conducted a Bayesian power analysis [@Kurz2019] by sampling from previous data [@Panizza2019] and simulating new datasets with varying sample sizes. We first fitted a Bayesian multilevel model to previous data, estimating payoff and normative utility component parameters for each participant (see Statistical Analysis). We then set the power analysis sample size $N$ and ran 100 simulations. For each simulation, we sampled $N$ participants with replacement from previous data, simulated normally-distributed mean SDO values that were correlated with the payoff utility component (_r_ = 0.24), and simulated normally-distributed mean RWA values that were correlated with the normative utility component (_r_ = 0.20). These effect sizes were estimated from previous work [@Claessens2020b; @Fischer_unpublished]. We then fitted a Bayesian multilevel model to this simulated dataset, and determined whether the 95% credible intervals for both interaction parameters (SDO interacting with the payoff utility component, and RWA interacting with the normative utility component) excluded zero. We used interaction effects to determine the correlations between the two components of norm-dependent utility and the two dimensions of political ideology (see Statistical Analysis). Our measure of statistical power was the proportion of 100 simulations in which the 95% credible intervals for both interaction parameters excluded zero. This power analysis suggested that, in order to achieve statistical power of 90%, we would require a sample of 450 participants.

### Participants and exclusions

We aimed to sample 470 participants (450 + 20 additional participants as a buffer for exclusions) on the online data collection platform Prolific (https://www.prolific.co/). Participants were from the United States and were required to have an approval rating of 95% or above on Prolific. In a single session of data collection on 12^th^ February 2021, we sampled `r nrow(ch6_d0)` participants. `r nrow(ch6_d0) - nrow(ch6_d)` of these participants were excluded for either (1) failing an attention check question, (2) flatlining on the self-report questions, or (3) completing the survey in less than 4 minutes (pre-registered exclusion criteria). This resulted in a final sample of `r nrow(ch6_d)` participants (`r sum(ch6_d$Gender == "Female")` female; age _M_ = `r printnum(mean(ch6_d$Age, na.rm = TRUE), digits = 0)` years, _SD_ = `r printnum(sd(ch6_d$Age, na.rm = TRUE), digits = 0)` years, range = `r printnum(min(ch6_d$Age, na.rm = TRUE), digits = 0)` - `r printnum(max(ch6_d$Age, na.rm = TRUE), digits = 0)` years).

### Materials

#### Mini-Dictator Games

Adapted from @Panizza2019, participants completed a series of 18 mini-DGs in which they chose between two possible distributions of 60 points between themselves and another participant. For example, in one mini-DG, dictators chose between (1) keeping 30 points for themselves and giving 30 points to the other participant, or (2) keeping all 60 points and giving nothing to the other participant. See Appendix to Chapter 6, Table \@ref(tab:ch6tableMiniDG) for the options in each of the 18 mini-DGs, taken from @Panizza2019. Each mini-DG appeared in a randomised order. Participants were truthfully told that, to determine payments after the study, one mini-DG would be randomly chosen for payment, all participants would be randomly matched into pairs, and one participant would be randomly selected as the dictator.

#### Norm Elicitation

Participants evaluated the social appropriateness of each possible Dictator Game split by completing the Krupka-Weber norm elicitation task, a measure of perceived social norms [@Krupka2013]. In this task, participants used a four-point Likert scale to rate the social appropriateness level that they thought _most other participants_ would choose for each possible mini-DG distribution. Social appropriateness ratings varied between "very socially inappropriate" and "very socially appropriate". Participants were truthfully told that, after the study, one mini-DG split would be chosen at random and, if their rating matched the majority in the session, they would be paid a bonus `r knitr::asis_output("\U00A3")`0.50.

#### Rule Following Task

This measure of rule following propensity was adapted from @Kimbrough2018. This task was included as a validity check: the normative utility component was shown to correlate with rule following in @Panizza2019. In the task, participants placed 30 balls in one of two buckets on the screen (side randomly counterbalanced). Participants earned 1 point per ball in one bucket, and 2 points per ball in the other bucket. However, participants were told that the "rule" was to place the balls in the bucket that earned them fewer points. The number of balls they placed in the bucket that earned them fewer points was used as a measure of rule following.

#### Questionnaire

Participants completed a questionnaire which measured social dominance orientation [@Sidanius2001], right-wing authoritarianism [@Altemeyer1996], and participants' age, gender, education, and political party support. See Appendix to Chapter 6, Table \@ref(tab:ch6tableSelfReport) for self-report items used in this study, and see pre-registration (https://osf.io/yk46b/) for a full list of measures included in the survey.

### Procedure

After agreeing to take part in the study on Prolific, participants were re-directed to Qualtrics (https://www.qualtrics.com/) to complete the online survey. Participants first read an information sheet that explained the purpose of the study, the process of de-identifying participants' study data, and participants' right to withdraw. Participants provided informed consent before beginning the study. To control for order effects, the Rule Following Task and Questionnaire blocks randomly appeared before or after the mini-DG and Norm Elicitation blocks. The Norm Elictation block always appeared directly after the mini-DGs. On average, participants took `r printnum(mean(pull(ch6_d[-436,"Duration (in seconds)"]) / 60), digits = 0) # 436 = outlier` minutes to complete the survey (_SD_ = `r printnum(sd(pull(ch6_d[-436,"Duration (in seconds)"]) / 60), digits = 0)` mins, range = `r printnum(min(pull(ch6_d[-436,"Duration (in seconds)"]) / 60), digits = 0)` - `r printnum(max(pull(ch6_d[-436,"Duration (in seconds)"]) / 60), digits = 0)` mins).

Participants earned a base rate of `r knitr::asis_output("\U00A3")`3.75 for taking part (all participants are paid in pounds sterling on Prolific). Throughout the survey, participants earned "points" which were converted to bonus payment at a rate of 1 point = `r knitr::asis_output("\U00A3")`0.02. Participants could earn a bonus of between `r knitr::asis_output("\U00A3")`0.00 - `r knitr::asis_output("\U00A3")`1.20 for the mini-DGs, between `r knitr::asis_output("\U00A3")`0.60 - `r knitr::asis_output("\U00A3")`1.20 for the Rule Following Task, possibly a bonus `r knitr::asis_output("\U00A3")`0.50 for the normative evaluations, and between `r knitr::asis_output("\U00A3")`0.00 - `r knitr::asis_output("\U00A3")`2.00 for other aspects of the study. On average, participants earned a bonus of `r knitr::asis_output("\U00A3")``r printnum(mean(ch6_d$bonus))` in total (_SD_ = `r knitr::asis_output("\U00A3")``r printnum(sd(ch6_d$bonus))`, range = `r knitr::asis_output("\U00A3")``r printnum(min(ch6_d$bonus))` - `r knitr::asis_output("\U00A3")``r printnum(max(ch6_d$bonus))`).

### Pre-registration

We pre-registered this study on the Open Science Framework (https://osf.io/yk46b/) on 11^th^ February 2021. We pre-registered three hypotheses. First, we hypothesised that we would recover the payoff and normative utility components from these data, replicating @Panizza2019 and supporting norm-dependent utility models from behavioural economics. Second, we hypothesised that SDO would be positively correlated with the payoff utility component. Third, we hypothesised that RWA would be positively correlated with the normative utility component.

### Statistical analysis

Norm-dependent utility models in behavioural economics assume that a participant $i$ playing a Dictator Game derives utility $U_{i}$ from both their own gains and from normative compliance [@Kessler2012]. This is captured by the utility function:

$$
U_{i}(\pi_{d},\pi_{r}) = \beta_{i}\pi_{d} + \gamma_{i}N_{i}(\pi_{d},\pi_{r})
$$

where $\pi_{d}$ is the payoff to the dictator (oneself), $\pi_{r}$ is the payoff to the receiver, $N_{i}$ is a norm function, and $\beta_{i}$ and $\gamma_{i}$ are the payoff and normative utility components, respectively. $\beta_{i}$ reflects the weight individuals give to their own gains. As $\beta_{i}$ increases, individuals behave more selfishly to maximise their personal gain. $\gamma_{i}$ reflects the weight individuals give to normative compliance. As $\gamma_{i}$ increases, individuals conform more to social norms.

Following @Panizza2019, we estimated $\beta_{i}$ and $\gamma_{i}$ in our data using a Bayesian multilevel logistic regression. We used logistic regression since the outcome variable is binary: whether participants chose the "selfish" option (i.e., the option that gives the dictator the higher payoff; $\text{non-selfish}_{i} = 0$) or the "non-selfish" option (i.e., the option that gives the dictator the lower payoff; $\text{non-selfish}_{i} = 1$) in each mini-DG choice. We opted for a logistic regression over a probit regression to be consistent with prior work [@Panizza2019].

To estimate $\beta_{i}$ and $\gamma_{i}$, we assume that the probability of choosing the option with the highest utility is proportional to the utility difference between the two options [@McFadden1976]. This results in the following logistic regression equation [@Panizza2019]:

$$
\begin{gathered}
\text{non-selfish}_{i}\sim\text{Bernoulli}(p_{i}) \\
\text{logit}(p_{i})=\alpha_{i}+\beta_{i}(\pi_{d}^{\text{nonself}}-\pi_{d}^{\text{self}})+\gamma_{i}(N_{i}^{\text{nonself}}-N_{i}^{\text{self}})
\end{gathered}
$$

where $\pi_{d}^{\text{nonself}}-\pi_{d}^{\text{self}}$ refers to the difference in dictator payoffs between the non-selfish and selfish options, and $N_{i}^{\text{nonself}}-N_{i}^{\text{self}}$ refers to the difference in social appropriateness ratings between the non-selfish and selfish options. In a Bayesian multilevel logistic regression model, we included $\alpha_{i}$ as a random intercept and $\beta_{i}$ and $\gamma_{i}$ as random slopes for participants. Since social appropriateness is rated on a four-point Likert scale, and thus the difference between social appropriateness ratings $N_{i}^{\text{nonself}}-N_{i}^{\text{self}}$ is ordinal, we modelled $\gamma$ as a monotonic effect [@Burkner2020]. 

To test our main hypotheses, we then determined whether the payoff and normative utility components were correlated with the two dimensions of political ideology. To achieve this in a multilevel framework, we allowed $\beta_{i}$ and $\gamma_{i}$ to interact with participant-level mean scores for SDO and RWA, respectively. For both interaction parameters, positive coefficients would indicate positive correlations between political ideology and the utility components, supporting our second and third pre-registered hypotheses.

We also conducted several exploratory analyses. Across all mini-DGs, we fitted a Bayesian multilevel logistic regression model with responses to the mini-DGs (selfish vs. non-selfish) as the binary outcome, random intercepts for participants and mini-DG items, and SDO and RWA as predictors with random slopes for mini-DG items. We also fitted a Bayesian multilevel cumulative regression model with Krupka-Weber normative evaluations as the ordinal outcome, random intercepts for participants and norm elicitation items, and SDO and RWA as predictors with random slopes for norm elicitation items. Finally, based on initial exploration of the data, we added an additional utility component ($\delta_{i}$) to the equation from @Panizza2019 to determine whether behaviour could be further explained by inequality aversion:

$$
\text{logit}(p_{i})=\alpha_{i}+\beta_{i}(\pi_{d}^{\text{nonself}}-\pi_{d}^{\text{self}})+\gamma_{i}(N_{i}^{\text{nonself}}-N_{i}^{\text{self}})+\delta_{i}(|\pi_{d}^{\text{self}}-\pi_{r}^{\text{self}}|-|\pi_{d}^{\text{nonself}}-\pi_{r}^{\text{nonself}}|)
$$

where $|\pi_{d}^{\text{self}}-\pi_{r}^{\text{self}}|$ reflects the absolute difference in payoffs between the dictator and receiver for the selfish option, and $|\pi_{d}^{\text{nonself}}-\pi_{r}^{\text{nonself}}|$ reflects the absolute difference in payoffs between the dictator and receiver for the non-selfish option. As the distribution of points between dictator and receiver becomes more unequal for the selfish option, the resulting value in the brackets increases. $\delta_{i}$ reflects inequality aversion. As $\delta_{i}$ increases, individuals express more aversion to inequality. In our exploratory analysis, we correlated this inequality aversion component with SDO.

All analyses were conducted in R version 4.0.2 [@R-base]. Bayesian models were fitted with the _brms_ package [@R-brms] and model fit was compared using leave-one-out cross-validation calculated with the _loo_ package [@R-loo]. For all models, we used weakly informative priors and the chains converged normally ($\hat{R}$ = 1). Figures were generated with the _ggplot2_ package [@R-ggplot2], and the reproducibility of all analyses was ensured using the _drake_ [@R-drake] package.

### Ethical approval

Ethical approval for this study was provided by the University of Auckland Human Participants Ethics Committee on 18^th^ January 2021 for three years (ref: UAHPEC21574).

## Results

Across all mini-DGs, there was substantial variation in the proportion of selfish vs. non-selfish decisions (Figure \@ref(fig:ch6plotCoopNorm)a). In particular, the highest proportion of non-selfish decisions came from the choice between splitting the points equally or keeping all the points to oneself (`r printnum((sum(ch6_d$DG10 == "30 points to you\n30 points to other participant") / nrow(ch6_d))*100, digits = 0)`% non-selfish decisions). This choice posed the starkest distinction between absolute equality and maximal inequality. By contrast, the lowest proportion of non-selfish decisions came from the choice between giving away all the points or keeping all the points to oneself; in this case, participants overwhelming chose to keep the points for themselves (`r printnum((sum(ch6_d$DG01 == "0 points to you\n60 points to other participant") / nrow(ch6_d))*100, digits = 0)`% non-selfish decisions). For this choice, both options resulted in maximally unequal distributions. In line with this mini-DG behaviour, the norm-elicitation task also revealed that participants believed an equal split of points was the most socially appropriate distribution (Figure \@ref(fig:ch6plotCoopNorm)b). As the distribution of points became more unequal, social appropriateness ratings declined, though keeping more than half of the points for oneself was deemed more inappropriate than giving more than half of the points away. These results show that participants are motivated by equality in these tasks.

(ref:ch6plotCoopNormCaption) _Behaviour in the mini-Dictator Game and norm elicitation tasks._ (_a_) The proportion of participants that chose the non-selfish option (i.e., the option that gives the dictator the lowest payoff) for each mini-DG decision. The x-axis states the payoff to the dictator, followed by the payoff to the other (i.e., self/other). (_b_) Social appropriateness ratings for different distributions of points in mini-DGs (1 = very socially inappropriate, 4 = very socially appropriate). Points are mean averages, lines represent +/- 2 standard errors.

```{r ch6plotCoopNorm, echo=FALSE, warning=F, message=F, fig.width=5, fig.height=8, fig.cap="(ref:ch6plotCoopNormCaption)"}
readd(ch6_plotCoopNorm)
```

To test our pre-registered hypotheses, we used participants' responses to the mini-DG and norm-elicitation tasks to estimate the payoff and normative utility components. Supporting our first hypothesis and replicating @Panizza2019, we found that participants' behaviour in the mini-DGs supported the existence of payoff and normative utility components. A model including payoff and normative utility parameters fitted the data better than the null intercept-only model (difference in expected log pointwise predictive density = `r printnum(abs(ch6_looDiff1[2,1]))`, _SE_ = `r printnum(ch6_looDiff1[2,2])`). In this model, the fixed effect parameter for the payoff utility component was positive ($\beta$ = `r printnum(ch6_summary1$fixed["diffP","Estimate"])`, 95% credible interval [`r printnum(ch6_summary1$fixed["diffP","l-95% CI"])` `r printnum(ch6_summary1$fixed["diffP","u-95% CI"])`]) indicating that participants were more likely to choose the non-selfish option as the difference in payoffs between the non-selfish and selfish options decreased (Appendix to Chapter 6, Figure \@ref(fig:ch6plotUtility)a). Another way of saying this is that participants were more likely to choose the selfish option as the relative payoffs of behaving selfishly increased, maximising personal gain. Similarly, the fixed effect parameter for the normative utility component was positive ($\gamma$ = `r printnum(ch6_summary1$fixed["modiffN","Estimate"])`, 95% CI [`r printnum(ch6_summary1$fixed["modiffN","l-95% CI"])` `r printnum(ch6_summary1$fixed["modiffN","u-95% CI"])`]) indicating that participants were more likely to choose the non-selfish option if the non-selfish option was rated as more socially appropriate than the selfish option (Appendix to Chapter 6, Figure \@ref(fig:ch6plotUtility)b). Model random effects revealed that the payoff and normative utility parameters varied substantially across individual participants (Appendix to Chapter 6, Figure \@ref(fig:ch6plotUtilityVar)). Replicating @Panizza2019, the frequency of rule following in the Rule Following Task was positively related to the normative utility component (_b_ = `r printnum(ch6_summary4$fixed["modiffN:followRule","Estimate"])`, 95% CI [`r printnum(ch6_summary4$fixed["modiffN:followRule","l-95% CI"])` `r printnum(ch6_summary4$fixed["modiffN:followRule","u-95% CI"])`]) but was unrelated to the payoff utility component (_b_ = `r printnum(ch6_summary11$fixed["diffP:followRule","Estimate"])`, 95% CI [`r printnum(ch6_summary11$fixed["diffP:followRule","l-95% CI"])` `r printnum(ch6_summary11$fixed["diffP:followRule","u-95% CI"])`]), suggesting that the normative utility component and the Rule Following Task are measuring similar psychological mechanisms.

However, contrary to our pre-registered hypotheses, the payoff and normative utility components were unrelated to SDO and RWA, respectively (Figure \@ref(fig:ch6plotSDORWA)). A model that interacted SDO with the payoff utility component and RWA with the normative utility component fitted the data slightly worse than the model without interaction effects (difference in expected log pointwise predictive density = `r printnum(ch6_looDiff2[2,1])`, _SE_ = `r printnum(ch6_looDiff2[2,2])`). The interaction model revealed that SDO was uncorrelated with the payoff utility component (_b_ = `r printnum(ch6_summary2a$fixed["diffP:sdo","Estimate"])`, 95% CI [`r printnum(ch6_summary2a$fixed["diffP:sdo","l-95% CI"])` `r printnum(ch6_summary2a$fixed["diffP:sdo","u-95% CI"])`]) and, while the effect was in the expected direction, RWA was uncorrelated with the normative utility component (_b_ = `r printnum(ch6_summary2a$fixed["modiffN:rwa","Estimate"])`, 95% CI [`r printnum(ch6_summary2a$fixed["modiffN:rwa","l-95% CI"])` `r printnum(ch6_summary2a$fixed["modiffN:rwa","u-95% CI"])`]). These results held when (1) interacting SDO and RWA with the payoff and normative utility components in separate models, (2) additionally controlling for age and gender, (3) using modal rather than individual social appropriateness ratings in the model, and (4) estimating the utility parameters from separate regressions for each participant and using these as outcome variables. Moreover, supporting the lack of relationship between RWA and the normative utility component, an exploratory analysis revealed that RWA was also unrelated to the probability of rule following in the rule following task (_b_ = `r printnum(ch6_summary9$fixed["rwa","Estimate"])`, 95% CI [`r printnum(ch6_summary9$fixed["rwa","l-95% CI"])` `r printnum(ch6_summary9$fixed["rwa","u-95% CI"])`]).

```{r ch6plotSDORWA, echo=FALSE, warning=F, message=F, fig.width=8, fig.height=4, fig.cap="\\textit{The relationships between the two utility components and the two dimensions of political ideology.} (\\textit{a}) Social dominance orientation is uncorrelated with the payoff utility component. (\\textit{b}) Right-wing authoritarianism is uncorrelated with the normative utility component. Lines are posterior median regression lines from Bayesian multilevel regression, shaded areas are 95% credible intervals."}
readd(ch6_plotSDORWA)
```

Despite the lack of support for our pre-registered hypotheses, further exploratory analyses revealed that SDO, but not RWA, was negatively related to cooperation in the mini-DGs. Participants lower in SDO were more likely to choose the non-selfish option than the selfish option across all games (_b_ = `r printnum(ch6_summary5$fixed["sdo","Estimate"])`, 95% CI [`r printnum(ch6_summary5$fixed["sdo","l-95% CI"])` `r printnum(ch6_summary5$fixed["sdo","u-95% CI"])`]; Figure \@ref(fig:ch6plotSDO)), whereas RWA did not predict the probability of selfishness (_b_ = `r printnum(ch6_summary5$fixed["rwa","Estimate"])`, 95% CI [`r printnum(ch6_summary5$fixed["rwa","l-95% CI"])` `r printnum(ch6_summary5$fixed["rwa","u-95% CI"])`]). Comparing the different mini-DGs, this model revealed that participants with the lowest SDO score were most likely to choose the non-selfish option when the choice was between an equal split and keeping all points for oneself (median probability of non-selfish choice for lowest SDO score = `r printnum(ch6_plotSDO$data %>% filter(sdo == min(.$sdo) & dg == "DG10") %>% pull(estimate__))`, 95% CI [`r printnum(ch6_plotSDO$data %>% filter(sdo == min(.$sdo) & dg == "DG10") %>% pull(lower__))` `r printnum(ch6_plotSDO$data %>% filter(sdo == min(.$sdo) & dg == "DG10") %>% pull(upper__))`]). These results suggest that equality is particularly important for participants low in SDO.

(ref:ch6plotSDOCaption) _The relationships between social dominance orientation and the probability of choosing the non-selfish option in each of the decisions in the mini-DG task, controlling for right-wing authoritarianism._ For each mini-DG option, the first number represents the amount of points kept by the dictator, and the second number represents the amount of points given to the other participant. Lines are posterior median regression lines from Bayesian multilevel regression, shaded areas are 95% credible intervals.

```{r ch6plotSDO, echo=FALSE, warning=F, message=F, fig.width=6, fig.height=6, fig.cap="(ref:ch6plotSDOCaption)"}
readd(ch6_plotSDO)
```

To further explore the relationship between SDO and preferences for equality, we added another term to the utility function which estimated the extent to which people weighted differences in equality between the two options into their decision-making (see Methods). Including inequality aversion in the utility function improved model fit above the model with only payoff and normative utility components (difference in expected log pointwise predictive density = `r printnum(abs(ch6_looDiff3[2,1]))`, _SE_ = `r printnum(ch6_looDiff3[2,2])`), suggesting that reducing inequality was an additional source of motivation for participants. In this model, the fixed effect parameter for the inequality aversion component was positive ($\delta$ = `r printnum(ch6_summary7$fixed["diffE","Estimate"])`, 95% CI [`r printnum(ch6_summary7$fixed["diffE","l-95% CI"])` `r printnum(ch6_summary7$fixed["diffE","u-95% CI"])`]) indicating that participants were more likely to choose the non-selfish option if the non-selfish option resulted in a more equal distribution of points (Appendix to Chapter 6, Figure \@ref(fig:ch6plotInequalityAversion)). As with the payoff and normative utility components, there was substantial variation in inequality aversion across participants (Appendix to Chapter 6, Figure \@ref(fig:ch6plotInequalityAversionVar)). When we additionally interacted SDO and RWA with the inequality aversion and normative utility parameters respectively, we found that the fixed interaction effect for SDO and inequality aversion was negative (_b_ = `r printnum(ch6_summary8$fixed["diffE:sdo","Estimate"])`, 95% CI [`r printnum(ch6_summary8$fixed["diffE:sdo","l-95% CI"])` `r printnum(ch6_summary8$fixed["diffE:sdo","u-95% CI"])`]) suggesting that participants with lower SDO scores were more averse to inequality than participants with higher SDO scores (Figure \@ref(fig:ch6plotInequalityAversionSDO)). In contrast, RWA continued to be unrelated to the normative utility component in this model (_b_ = `r printnum(ch6_summary8$fixed["modiffN:rwa","Estimate"])`, 95% CI [`r printnum(ch6_summary8$fixed["modiffN:rwa","l-95% CI"])` `r printnum(ch6_summary8$fixed["modiffN:rwa","u-95% CI"])`]).

(ref:ch6plotInequalityAversionSDOCaption) _Social dominance orientation is negatively related to inequality aversion._ Lines are posterior median regression lines from Bayesian multilevel regression, shaded areas are 95% credible intervals.

```{r echo=F, results='hide'}
# free up memory
rm(ch6_d0, ch6_d, ch6_looDiff1, ch6_looDiff2, ch6_looDiff3, 
   ch6_post5, ch6_plotSDO, ch6_plotRWA, ch6_summary1, ch6_summary2a,
   ch6_summary4, ch6_summary5, ch6_summary8, ch6_summary9, ch6_summary11)
gc()
```

```{r ch6plotInequalityAversionSDO, echo=FALSE, warning=F, message=F, fig.width=4, fig.height=4, fig.cap="(ref:ch6plotInequalityAversionSDOCaption)"}
# plot
readd(ch6_plotInequalityAversionSDO)
```

Exploratory analyses also revealed that RWA, but not SDO, was related to perceptions of the severity of social norms. In the norm-elicitation task, we found that participants higher in RWA were more likely to rate distributions of points as socially inappropriate (_b_ = `r printnum(ch6_summary3$fixed["rwa","Estimate"])`, 95% CI [`r printnum(ch6_summary3$fixed["rwa","l-95% CI"])` `r printnum(ch6_summary3$fixed["rwa","u-95% CI"])`]; Figure \@ref(fig:ch6plotRWA)), whereas SDO did not predict normative ratings (_b_ = `r printnum(ch6_summary3$fixed["sdo","Estimate"])`, 95% CI [`r printnum(ch6_summary3$fixed["sdo","l-95% CI"])` `r printnum(ch6_summary3$fixed["sdo","u-95% CI"])`]). The slope for RWA is most negative for distributions of points that favour the other participant (e.g., giving all points to the other), suggesting that participants with higher RWA scores believe that it is particularly socially inappropriate to give away too many resources.

(ref:ch6plotRWACaption) _The relationships between right-wing authoritarianism and the social appropriateness ratings given for each distribution of points in the norm elicitation task, controlling for social dominance orientation._ Lines are posterior median regression lines from Bayesian multilevel regression, shaded areas are 95% credible intervals.

```{r ch6plotRWA, echo=FALSE, warning=F, message=F, fig.width=6, fig.height=6, fig.cap="(ref:ch6plotRWACaption)"}
readd(ch6_plotRWA)
```

## Discussion

This study expanded upon previous research by relating the two dimensions of political ideology to behaviour in a novel task. The mini-Dictator Game paradigm addressed limitations of previous work by explicitly measuring social norms and attempting to differentiate between selfish and conformist preferences. In the task, we found that participants lower in SDO, but not RWA, were more cooperative, a preference which was driven specifically by inequality aversion. Participants higher in RWA, but not SDO, perceived norm violations to be more socially inappropriate. However, in contrast to our pre-registered hypotheses, the two components of norm-dependent utility (sensitivity to personal gain and sensitivity to normative compliance) were unrelated to the two dimensions of political ideology. These results have implications for both norm-dependent utility models in behavioural economics and the dual evolutionary foundations of political ideology.

We found that participants lower in SDO were more cooperative across all the mini-DGs. However, in contrast to our pre-registered hypothesis, we did not find that this cooperation from participants low in SDO was driven by giving less weight to personal gain. Instead, our exploratory analysis revealed that this cooperation was driven specifically by inequality aversion. This finding provides a psychological explanation for the well-supported relationship between SDO and cooperation [@Thielmann2020], suggesting that people low in SDO cooperate with others as a means to equalise payoffs. More generally, the finding also broadens the scope of social dominance theory [@Sidanius2001], suggesting that people low in SDO have a _general_ preference for equality that applies not just for social groups in the real world, but also for distributions of experimental points between individual strangers in abstract tasks.

The link between SDO and inequality aversion, rather than pure self-interest, further develops the dual evolutionary foundations theory of political ideology. This theory characterises the economic dimension of political ideology as reflecting behavioural preferences for cooperation [@Claessens2020a]. Under this view, concerns about equality could be understood as a mere side-effect of a general behavioural preference for cooperation. On the contrary, we have revealed that equality is a _primary psychological motivator_ for the cooperative behaviour underlying the economic dimension of political ideology: people lower in SDO cooperate _in order to_ equalise payoffs. This can help us understand the policy views of people low in SDO and economic progressives. For example, when economic progressives support higher taxes to fund free education and universal healthcare, their cooperation is likely psychologically motivated by a desire to bring about equality in the access to these services.

We found that participants high in RWA interpreted social norms as being harsher (i.e., authoritarians rated norm violations as more socially inappropriate), but they were no more likely to follow those norms as a result. This finding was further supported by the lack of relationship between RWA and rule following, a result we also found in a previous study of New Zealanders [@Claessens2020b]. Why does an authoritarian interpretation of harsh social norms not translate into subsequent adherence to those norms? One reason may be that in both the mini-DG and rule following tasks, decisions were made privately. If decisions had instead been made public to other participants, allowing the possibility of peer disapproval, authoritarians may have been more motivated to adhere to social norms. This could also explain why, in a previous study [@Claessens2020b], authoritarians conformed to social information in a visual estimation task: social information is not affected by whether or not someone else is watching. Another explanation for the lack of norm adherence is that, in our tasks, the social norms were not attached to particular socio-cultural groups. Authoritarians are predicted to adhere to the social norms of _their_ socio-cultural group [@Claessens2020a]. But in the mini-DG paradigm, social norms were estimated based on what "all other participants in the study" believed. This setup can explain our results: authoritarians may be predisposed to perceive social norms are harsh, but without cues linking particular social norms to particular groups, they cannot know which norms to adhere to. Future work could combine our methodology with minimal-group paradigms to determine if these cues affect norm adherence in authoritarians.

We also found that authoritarians were particularly disapproving of (or, more specifically, perceived there to be harsh social norms against) ultra-cooperative decisions in which more than half of one's resources were given away. This echoes previous work showing that RWA is positively associated with anti-social punishment: removing resources from generous individuals at a cost to oneself [@Claessens2020b]. Why might authoritarians disapprove of ultra-cooperative decisions? In psychology, negative reactions to this kind of behaviour are referred to as "do-gooder derogation" [@Minson2012; @Pleasant2018], triggered by perceived moral superiority from the giver and, possibly, moral judgement towards those who did not give. Since authoritarians are uniquely sensitive to normative condemnation, it makes sense that they would display these negative reactions to ultra-cooperative behaviour. This underexplored aspect of authoritarian psychology could potentially explain their opposition to welfare [@Blanchet2021] and environmentalism [@Stanley2019], insofar as these policies are seen as ultra-cooperative.

More generally, our results replicate previous work in behavioural economics showing that cooperative behaviour in economic games is motivated by adherence to social norms of fairness [@Krupka2013]. As in @Panizza2019, people tended to rate the equal distribution of points as the most socially appropriate, with social appropriateness ratings declining as distributions became more unequal. People's social appropriateness ratings also predicted their cooperative decision-making, supporting norm-dependent utility models in behavioural economics [@Kimbrough2018; @Krupka2009; @Krupka2013; @Kessler2012]. However, the addition of another utility parameter (inequality aversion) to the utility function improved model fit, suggesting that norm-dependent utility models alone are incomplete: people independently weigh both social norms and inequality when deciding how to distribute resources. Since social norms largely tracked equality in this study, the normative and inequality aversion utility parameters were positively correlated (_r_ = `r printnum(ch6_summary7$random$id["cor(diffE,modiffN)","Estimate"])`). But it will be interesting to study situations in which social norms do not coincide with equality, such as the norm of self-interest in market interactions [@Miller1999], to see how the two utility components influence behaviour when they are in conflict.

In sum, in line with the dual evolutionary foundations of political ideology [@Claessens2020a], we found that individual differences in cooperative and conformist preferences covaried with the two dimensions of political ideology. SDO was related to cooperation and inequality aversion, and RWA was related to perceptions of the severity of social norms. By contrast, we found that general selfishness and adherence to social norms were unrelated to political ideology. These findings suggest that inequality aversion and perceptions of norm severity, rather than selfishness and norm adherence, are the psychological motivations that underlie the behavioural variation in cooperation and conformity across the political landscape.
